### 2014-02-25 phpmyadmin/GSoC_infosession_05Mar2014/presentation.tex
update date and stats

\documentclass[14pt]{beamer}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usetheme{Madrid}
\definecolor{PmaPurple}{RGB}{00,33,66}
\definecolor{PmaOrange}{RGB}{255,173,17}
\definecolor{PmaOlive}{HTML}{727E0A}
\usecolortheme[named=PmaPurple]{structure}
\setbeamercolor{title}{fg=PmaOrange}
\setbeamercolor{frametitle}{fg=PmaOrange}
\hypersetup{colorlinks,linkcolor=,urlcolor=PmaOrange}
\logo{
  \includegraphics[scale=.6]{pictures/pma_logo.png}
}
\title{GSoC 2014 info session - phpMyAdmin}
\author{Dieter Adriaenssens}
\institute[phpMyAdmin]{phpMyAdmin GSoC mentor 2011-2013}
\date[5Mar2014]{Google Summer of Code info session UGent\\
Mentoring organisation : phpMyAdmin\\
March 5th, 2014}
\subject{phpMyAdmin}
\begin{document}
  \begin{frame}
  \titlepage
  \end{frame}
  \begin{frame}
    \frametitle{What is phpMyAdmin?}
    %(question to audience : who knows/uses pma?)
      A webinterface written in PHP (and JavaScript),\\
      to administer a MySQL database and users.
  \end{frame}
  \begin{frame}
    \frametitle{Current version}
    % (questions to audience about used version)
    \begin{itemize}
      \item stable version : 4.1.8
      \item in development : 4.2
      \item available in \href{http://www.phpmyadmin.net/home\_page/translations.php}{72 languages}
      \item avg. 100K downloads (every release)
      \item 3M downloads in total last year
      \item 10K commits by 285 contributors last year
      \item approx. 20 active contributors every month
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{Why phpMyAdmin participates in GSoC}
    \begin{itemize}
      \item to get new features implemented in future versions
      \item attract new developers
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{GSoC history}
    \begin{itemize}
      \item 2012 : 6 students, 5 passed -> phpMyAdmin 4.0
      \item 2011 : 8 students, 7 passed -> phpMyAdmin 3.5
      \item 2010 : 6 students, 6 passed -> phpMyAdmin 3.4
      \item 2009 : under MySQL umbrella : 5 students
      \item 2008 : under MySQL umbrella : 2 students
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{What made it into 3.5}
    \begin{itemize}
      \item improved support for stored routines, events and triggers
      \item openGIS support
      \item Drizzle support
      \item improved ENUM/SET editor
      \item advisor : suggestions to tweak config setting in order to optimise MySQL server performance
      \item zoom search
      \item grid editing
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{What made it into 4.0}
    \begin{itemize}
      \item code cleanup and refactoring
      \item plugin system for import/export/auth/\ldots
      \item removal of frames
      \item new navigation tree
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{Ideas list for GSoC 2013}
    Full list : \href{http://wiki.phpma.org/pma/GSoC\_2013\_Ideas\_List}{wiki.phpma.org/pma/GSoC\_2013\_Ideas\_List}
    \begin{itemize}
      \item automated testing
      \item refactoring
      \item synchronize data
      \item AJAX error reporting
      \item interface improvements
      \item your idea\ldots
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{What to expect}
    \begin{itemize}
      \item mentoring by experienced mentors/developers
      \item be active on the mailinglist
      \item submit a patch fixing a bug or adding a small feature
      \item 12 weeks of 40 hours coding
    \end{itemize}
 \end{frame}
  \begin{frame}
    \frametitle{Submitting a good proposal}
    \begin{itemize}
      \item write a good project proposal :
      \begin{itemize}
        \item detailed functional description
        \item well defined deliverables
        \item clear schedule
      \end{itemize}
      \item submit well in advance : time for feedback
      \item application deadline : May 3
      \item patch deadline (phpMyAdmin) : May 10
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{During GSoC}
    \textbf{May 27} - When you get selected :
    \begin{itemize}
      \item community bonding period : get up to speed, read, ask questions, set up your wiki, clone the git repo, Google administration
      \pause
      \item \textbf{June 17} - start coding, first payment
      \item mandatory weekly reports on your blog, commit code regularly, merged into master by mentor
      \pause
      \item \textbf{August 2} - mid-term evaluation, based on your schedule + payment
      \pause
      \item finish coding, documentation, testing
      \pause
      \item \textbf{September 16} - pencils down, final evaluation, submit code, final payment
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{What happens after GSoC}
    \begin{itemize}
      \item your code gets merged with master branch
      \item we offer active membership :
      \begin{itemize}
        \item commit access
        \item team membership
        \item invitation to team meetings
      \end{itemize}
    \end{itemize}
  \end{frame}
  \begin{frame}
   \frametitle{Questions}
    Thanks for your attention!\\
    Questions?
    \begin{itemize}
      \item website : \href{http://www.phpmyadmin.net/}{www.phpmyadmin.net}
      \item GSoC info : \href{http://wiki.phpma.org/pma/GSoC\_2013\_Applicant\_Guide}{wiki.phpma.org/pma/GSoC\_2013\_Applicant\_Guide}
      \item developers mailing list (development discussion): \href{mailto:phpmyadmin-devel@lists.sourceforge.net}{phpmyadmin-devel@lists.sourceforge.net}
    \end{itemize}
  \end{frame}
\end{document}

### 2014-02-26 phpmyadmin/GSoC_infosession_05Mar2014/presentation.tex
update links, dates and ideas to 2014

@@
     \begin{itemize}
-      \item to get new features implemented in future versions
+      \item get new features implemented in future versions
       \item attract new developers
@@
     \begin{itemize}
+      \item 2013 : 6 students, 6 passed -> phpMyAdmin 4.1
       \item 2012 : 6 students, 5 passed -> phpMyAdmin 4.0
@@
   \begin{frame}
-    \frametitle{Ideas list for GSoC 2013}
-    Full list : \href{http://wiki.phpma.org/pma/GSoC\_2013\_Ideas\_List}{wiki.phpma.org/pma/GSoC\_2013\_Ideas\_List}
+    \frametitle{Ideas list for GSoC 2014}
+    Full list : \href{http://wiki.phpma.org/pma/GSoC\_2014\_Ideas\_List}{wiki.phpma.org/pma/GSoC\_2014\_Ideas\_List}
     \begin{itemize}
       \item automated testing
-      \item refactoring
-      \item synchronize data
-      \item AJAX error reporting
-      \item interface improvements
+      \item (user) interface improvements
+      \item Error reporting
+      \item structure tools 
+      \begin{itemize}
+        \item database normalization
+        \item central list of columns
+      \end{itemize}
       \item your idea\ldots
@@
       \item submit well in advance : time for feedback
-      \item application deadline : May 3
-      \item patch deadline (phpMyAdmin) : May 10
+      \item \textbf{March 21st} : application deadline
+      \item \textbf{March 28th} : patch deadline (phpMyAdmin)
     \end{itemize}
@@
     \frametitle{During GSoC}
-    \textbf{May 27} - When you get selected :
+    When you get selected :
     \begin{itemize}
-      \item community bonding period : get up to speed, read, ask questions, set up your wiki, clone the git repo, Google administration
-      \pause
-      \item \textbf{June 17} - start coding, first payment
+      \item community bonding period : get up to speed, read, ask questions, set up your wiki, clone the git repo, Google 
       \item mandatory weekly reports on your blog, commit code regularly, merged into master by mentor
-      \pause
-      \item \textbf{August 2} - mid-term evaluation, based on your schedule + payment
-      \pause
-      \item finish coding, documentation, testing
-      \pause
-      \item \textbf{September 16} - pencils down, final evaluation, submit code, final payment
     \end{itemize}
@@
       \item website : \href{http://www.phpmyadmin.net/}{www.phpmyadmin.net}
-      \item GSoC info : \href{http://wiki.phpma.org/pma/GSoC\_2013\_Applicant\_Guide}{wiki.phpma.org/pma/GSoC\_2013\_Applicant\_Guide}
+      \item GSoC info : \href{http://wiki.phpma.org/pma/GSoC\_2014\_Applicant\_Guide}{wiki.phpma.org/pma/GSoC\_2014\_Applicant\_Guide}
       \item developers mailing list (development discussion): \href{mailto:phpmyadmin-devel@lists.sourceforge.net}{phpmyadmin-devel@lists.sourceforge.net}

### 2014-02-26 phpmyadmin/GSoC_infosession_05Mar2014/presentation.tex
update previous project list

@@
   \begin{frame}
-    \frametitle{What made it into 3.5}
-    \begin{itemize}
-      \item improved support for stored routines, events and triggers
-      \item openGIS support
-      \item Drizzle support
-      \item improved ENUM/SET editor
-      \item advisor : suggestions to tweak config setting in order to optimise MySQL server performance
-      \item zoom search
-      \item grid editing
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{What made it into 4.0}
+    \frametitle{Successful projects in previous editions}
     \begin{itemize}
@@
       \item plugin system for import/export/auth/\ldots
+      \item openGIS support
+      \item Drizzle support
+      \item advisor to optimise MySQL server performance
+      \item interface improvements : zoom search, grid editing, new navigation tree, \ldots
       \item removal of frames
-      \item new navigation tree
+      \item AJAX error reporting 
     \end{itemize}
@@
     \frametitle{Ideas list for GSoC 2014}
-    Full list : \href{http://wiki.phpma.org/pma/GSoC\_2014\_Ideas\_List}{wiki.phpma.org/pma/GSoC\_2014\_Ideas\_List}
+    Full list :
+    \href{http://wiki.phpmyadmin.net/pma/GSoC\_2014\_Ideas\_List}{wiki.phpmyadmin.net/pma/GSoC\_2014\_Ideas\_List}
     \begin{itemize}
-      \item automated testing
-      \item (user) interface improvements
-      \item Error reporting
       \item structure tools 
@@
       \end{itemize}
+      \item error reporting improvements
+      \item (user) interface improvements
+      \item automated testing
       \item your idea\ldots
@@
   \begin{frame}
-    \frametitle{What happens after GSoC}
+    \frametitle{After GSoC}
     \begin{itemize}

### 2014-03-24 iptables_newline2014/presentation.tex
create presentation about iptables

\documentclass[14pt]{beamer}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usetheme{Madrid}
%\usecolortheme[named=PmaPurple]{structure}
%\setbeamercolor{title}{fg=PmaOrange}
%\setbeamercolor{frametitle}{fg=PmaOrange}
%\hypersetup{colorlinks,linkcolor=,urlcolor=PmaOrange}
\title[Reduce iptables config complexity]{Reducing iptables configuration complexity using chains}
\author{Dieter Adriaenssens}
%\institute[phpMyAdmin]{phpMyAdmin developer}
\date[29Mar2014]{Newline 0x04\\
Whitespace, Ghent\\
March 29th, 2014}
\subject{iptables}
\begin{document}
  \begin{frame}
  \titlepage
  \end{frame}
  \begin{frame}
    \frametitle{What is iptables}
    %(question to audience : who knows/uses iptables?)
  \end{frame}
  \begin{frame}
   \frametitle{Questions}
    Thanks for your attention!\\
    Questions?\\
    Contact
    \begin{itemize}
      \item Blog : \href{http://ruleant.blogspot.com/}{http://ruleant.blogspot.com}
      \item Twitter : \href{https://twitter.com/dcadriaenssens}{@dcadriaenssens}
    \end{itemize}
  \end{frame}
\end{document}

### 2014-03-26 iptables_newline2014/presentation.tex
add slides explaining iptables

@@
 \usetheme{Madrid}
-%\usecolortheme[named=PmaPurple]{structure}
-%\setbeamercolor{title}{fg=PmaOrange}
-%\setbeamercolor{frametitle}{fg=PmaOrange}
-%\hypersetup{colorlinks,linkcolor=,urlcolor=PmaOrange}
 \title[Reduce iptables config complexity]{Reducing iptables configuration complexity using chains}
 \author{Dieter Adriaenssens}
-%\institute[phpMyAdmin]{phpMyAdmin developer}
 \date[29Mar2014]{Newline 0x04\\
@@
   \begin{frame}
-    \frametitle{What is iptables}
+    \frametitle{What is iptables?}
     %(question to audience : who knows/uses iptables?)
+    iptables is a tool for creating the rulesets for netfilter, a packet filtering framework which was introduced in the linux 2.4 kernel
+  \end{frame}
+  \begin{frame}
+    \frametitle{Rules}
+    When an IP packet comes in, it is checked against a set of rules.
+    \begin{example}
+      \small{iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport http -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
+      iptables -A INPUT -j REJECT}
+    \end{example}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Chains}
+    Predefined chains :
+    \begin{itemize}
+      \item INPUT
+      \item OUTPUT
+      \item FORWARD
+      \item PREROUTING
+      \item POSTROUTING
+    \end{itemize}
+    You can define your own chains
+  \end{frame}
+  \begin{frame}
+    \frametitle{Filters}
+    Filter on packet parameters :
+    \begin{itemize}
+      \item protocol (tcp, udp, icmp, \ldots)
+      \item destination/source port
+      \item destination/source IP address
+      \item in/outgoing interface (eth0, \ldots)
+      \item \ldots
+    \end{itemize}
+    \begin{example}
+      \small{iptables -A INPUT -p tcp --dport ssh -s 10.0.0.0/8 -j ACCEPT}
+    \end{example}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Targets}
+    What to do if a packet matches a rule :
+    \begin{itemize}
+      \item ACCEPT
+      \item DROP
+      \item QUEUE $\rightarrow$ userspace
+      \item RETURN $\rightarrow$ leave current chain
+      \item LOG
+      \item jump to a custom chain
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Putting it all together}
+    \begin{example}
+      \small{iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
+      iptables -A INPUT -j REJECT}
+    \end{example}
+    But you can define default behaviour for a predefined chain:
+    \begin{example}
+      \small{iptables -P INPUT DROP}
+    \end{example}
   \end{frame}

### 2014-03-27 iptables_newline2014/presentation.tex
add section about default target and established state

@@
 \usepackage{hyperref}
+\usepackage{ulem}
 \usepackage[T1]{fontenc}
@@
       iptables -A INPUT -p tcp --dport https -j ACCEPT\\
-      iptables -A INPUT -j REJECT}
+      iptables -A INPUT -j DROP}
     \end{example}
@@
     \frametitle{Putting it all together}
+    All rules are checked one for one, and if one matches the target is executed :
     \begin{example}
@@
       iptables -A INPUT -p tcp --dport https -j ACCEPT\\
-      iptables -A INPUT -j REJECT}
+      iptables -A INPUT -j DROP}
     \end{example}
-    But you can define default behaviour for a predefined chain:
+    The last rule matches everything, so if a packet didn't match a previous rule, it will be rejected.
+  \end{frame}
+  \begin{frame}
+    \frametitle{Default target DROP}
+    But you can define default target for the predefined chains:
     \begin{example}
-      \small{iptables -P INPUT DROP}
+      \small{iptables -P INPUT DROP\\
+      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
+      \sout{iptables -A INPUT -j DROP}}
     \end{example}
+    If none of the rules match, the default target is executed.\\
+    In this example, a packet is dropped.
+  \end{frame}
+  \begin{frame}
+    \frametitle{Default target ACCEPT}
+    \textbf{Exercise:}\\
+    What happens if the default target is ACCEPT?
+    \begin{example}
+      \small{iptables -P INPUT ACCEPT\\
+      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport https -j ACCEPT}
+    \end{example}
+    \pause
+    All packets are accepted, even if no rule matches them.\\
+    \pause
+    \textbf{Always define a default target or use an all-matching rule at the end!}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Firewall for your server}
+    Filter both incoming and outgoing traffic :
+    \begin{example}
+      \small{iptables -P INPUT DROP\\
+      iptables -P OUTPUT DROP\\
+      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport https -j ACCEPT}
+    \end{example}
+    \pause
+    \textbf{Problem:} incoming packets are accepted, but replies are dropped!
+  \end{frame}
+  \begin{frame}
+    \frametitle{Firewall for your server}
+    Also add rules for outgoing traffic :
+    \begin{example}
+      \small{iptables -P INPUT DROP\\
+      iptables -P OUTPUT DROP\\
+      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
+      iptables -A OUTPUT -p tcp --sport ssh -j ACCEPT\\
+      iptables -A OUTPUT -p tcp --sport 80 -j ACCEPT\\
+      iptables -A OUTPUT -p tcp --sport https -j ACCEPT}
+    \end{example}
+    \pause
+    This can get complicated if more rules are added.
+  \end{frame}
+  \begin{frame}
+    \frametitle{Established state}
+    \textbf{Solution:} check for established state
+    \begin{example}
+      \small{iptables -P INPUT DROP\\
+      iptables -P OUTPUT DROP\\
+      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
+      iptables -A OUTPUT -m state --state ESTABLISHED -j ACCEPT}
+    \end{example}
+    This makes the rules more manageable, especially when output rules are also defined.
   \end{frame}

### 2014-03-27 iptables_newline2014/presentation.tex
add match parameter

@@
     \begin{example}
-      \small{iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport http -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
+      \small{iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport http -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
       iptables -A INPUT -j DROP}
@@
     \begin{example}
-      \small{iptables -A INPUT -p tcp --dport ssh -s 10.0.0.0/8 -j ACCEPT}
+      \small{iptables -A INPUT -m tcp -p tcp --dport ssh -s 10.0.0.0/8 -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
+      \small{iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
       iptables -A INPUT -j DROP}
@@
       \small{iptables -P INPUT DROP\\
-      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
       \sout{iptables -A INPUT -j DROP}}
@@
       \small{iptables -P INPUT ACCEPT\\
-      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport https -j ACCEPT}
+      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
     \end{example}
@@
       iptables -P OUTPUT DROP\\
-      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport https -j ACCEPT}
+      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
     \end{example}
@@
       iptables -P OUTPUT DROP\\
-      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
-      iptables -A OUTPUT -p tcp --sport ssh -j ACCEPT\\
-      iptables -A OUTPUT -p tcp --sport 80 -j ACCEPT\\
-      iptables -A OUTPUT -p tcp --sport https -j ACCEPT}
+      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      iptables -A OUTPUT -m tcp -p tcp --sport ssh -j ACCEPT\\
+      iptables -A OUTPUT -m tcp -p tcp --sport 80 -j ACCEPT\\
+      iptables -A OUTPUT -m tcp -p tcp --sport https -j ACCEPT}
     \end{example}
@@
       iptables -P OUTPUT DROP\\
-      iptables -A INPUT -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -p tcp --dport https -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
       iptables -A OUTPUT -m state --state ESTABLISHED -j ACCEPT}

### 2014-03-27 iptables_newline2014/presentation.tex
add RELATED state
add setup and summary

@@
       iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      iptables -A OUTPUT -m state --state ESTABLISHED -j ACCEPT}
+      iptables -A OUTPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}
     \end{example}
     This makes the rules more manageable, especially when output rules are also defined.
+  \end{frame}
+  \begin{frame}
+    \frametitle{Realworld example : setup}
+    \begin{itemize}
+      \item Tcp port 80 (http) and 443 (https) is available for all.
+      \item ssh (port 20) and webmin (port 10000), is limited to admin PCs.
+      \item A SMB service is limited to admin and webmaster PCs.
+      \item Some outgoing connections are allowed (dns, dhcp, smtp, ntp).
+      \item Some external websites are reachable.
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Summary}
+    \begin{itemize}
+      \item use ESTABLISHED state to reduce number of rules
+      \item using chains makes your rules easier readable and maintainable
+      \item chains can be reused for several rules
+      \item chains can be chained together
+      \item faster, because it only jumps to a chain when a rule matches
+    \end{itemize}
   \end{frame}

### 2014-03-28 iptables_newline2014/presentation.tex
add slides for webserver and ssh/webmin only accessible for Admin PCs

@@
       \item Tcp port 80 (http) and 443 (https) is available for all.
-      \item ssh (port 20) and webmin (port 10000), is limited to admin PCs.
+      \item SSH (port 22) and a webbased admin tool (port 10000), is limited to admin PCs.
       \item A SMB service is limited to admin and webmaster PCs.
@@
   \end{frame}
+  \begin{frame}
+    \frametitle{Realworld example : webserver}
+    Webservice, tcp port 80 (http) and 443 (https) is available for all :
+    \begin{example}
+      \small{iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
+    \end{example}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Realworld example : sysadmin PCs}
+    Sysadmins are allowed to access SSH (tcp port 22) and a webbased admin tool (tcp port 10000).\\
+    The IP addresses of their PCs :
+    \begin{itemize}
+	\item 10.100.2.3
+	\item 10.100.2.4
+	\item 10.100.2.7
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Realworld example : sysadmin PCs}
+    \begin{example}
+      \small{iptables -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 22 -j ACCEPT\\
+      iptables -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 22 -j ACCEPT\\
+      iptables -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 22 -j ACCEPT\\
+      iptables -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 10000 -j ACCEPT\\
+      iptables -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 10000 -j ACCEPT\\
+      iptables -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 10000 -j ACCEPT}
+    \end{example}
+    \pause
+    The same IP addresses are repeated. If there is a change, several rules have to be updated.\\
+    \pause
+    \textbf{This can be done more efficient!}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Realworld example : sysadmin PCs}
+    Wouldn't it be convenient to :
+    \begin{itemize}
+      \item make a list of IP addresses?
+      \item check this list when a packet matches a rule (fe. TCP port 22)?
+      \item reuse this list when another rule matches a packet?
+    \end{itemize}
+    \pause
+    \textbf{This is possible with chains!}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Introducing chains}
+    Firstly, create the chain and add the rules
+    \begin{example}
+      \small{iptables -N admin\_IP\\
+      iptables -A admin\_IP -s 10.100.2.3 -j ACCEPT\\
+      iptables -A admin\_IP -s 10.100.2.4 -j ACCEPT\\
+      iptables -A admin\_IP -s 10.100.2.7 -j ACCEPT\\
+      iptables -A admin\_IP -j DROP}
+    \end{example}
+    \pause
+    \begin{itemize}[<+->]
+      \item Syntax of adding rules to a custom chain is similar to adding to default targets.
+      \item Custom chains don't have a default target, so set a target for all packets that are not matched.
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Realworld example : sysadmin PCs}
+    Add the rules for TCP port 22 and 10000 using the custom chain :
+    \begin{example}
+      \small{iptables -A INPUT -p tcp -m tcp --dport 22 -j admin\_IP\\
+      iptables -A INPUT -p tcp -m tcp --dport 10000 -j admin\_IP}
+    \end{example}
+  \end{frame}
+
   \begin{frame}

### 2014-03-28 iptables_newline2014/presentation.tex
add Benefits of using chains
add chaining chains

@@
   \end{frame}
+  \begin{frame}
+    \frametitle{Benefits}
+    \begin{itemize}
+      \item the list of IP addresses in the custom chain is reused for both ports, so they have to be defined only once.
+      \item adding/changing/removing an IP address is much easier
+      \item a better overview of the firewall rules.
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Realworld example : webmaster PCs}
+    Let's do the same for webmasters AND admins having access to SMB.\\
+    Create a chain with webmaster IP addresses :
+    \begin{example}
+      \small{iptables -N webmaster\_IP\\
+      iptables -A webmaster\_IP -s 10.100.2.11 -j ACCEPT\\
+      iptables -A webmaster\_IP -s 10.100.2.17 -j ACCEPT\\
+      iptables -A webmaster\_IP -s 10.100.2.34 -j ACCEPT\\
+      iptables -A webmaster\_IP -j DROP}
+    \end{example}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Realworld example : access to SMB}
+    Add the rules for SMB (tcp port 139 and 445) using the custom chains :
+    \begin{example}
+      \small{iptables -A INPUT -p tcp -m tcp --dport 445 -j admin\_IP\\
+      iptables -A INPUT -p tcp -m tcp --dport 445 -j webmaster\_IP\\
+      iptables -A INPUT -p tcp -m tcp --dport 139 -j admin\_IP\\
+      iptables -A INPUT -p tcp -m tcp --dport 139 -j webmaster\_IP}
+    \end{example}
+  \pause
+  \textbf{This can be optimized!}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Chaining chains}
+    Include the sysadmin chain in the webmaster chain:
+    \begin{example}
+      \small{iptables -N webmaster\_IP\\
+      iptables -A webmaster\_IP -s 10.100.2.11 -j ACCEPT\\
+      iptables -A webmaster\_IP -s 10.100.2.17 -j ACCEPT\\
+      iptables -A webmaster\_IP -s 10.100.2.34 -j ACCEPT\\
+      \sout{iptables -A webmaster\_IP -j DROP}\\
+      iptables -A webmaster\_IP -j admin\_IP}
+    \end{example}
+    Jump to the sysadmin chain at the end of the webmaster chain.\\
+    \textbf{Warning :} check conditions of chain ending
+  \end{frame}
+  \begin{frame}
+    \frametitle{Realworld example : access to SMB}
+    The rules for SMB (tcp port 139 and 445) can be reduced to :
+    \begin{example}
+      \small{\sout{iptables -A INPUT -p tcp -m tcp --dport 445 -j admin\_IP}\\
+      iptables -A INPUT -p tcp -m tcp --dport 445 -j webmaster\_IP\\
+      \sout{iptables -A INPUT -p tcp -m tcp --dport 139 -j admin\_IP}\\
+      iptables -A INPUT -p tcp -m tcp --dport 139 -j webmaster\_IP}
+    \end{example}
+  The admin chain is checked as well because the webmaster chain includes it.
+  \end{frame}
 
   \begin{frame}
-    \frametitle{Summary}
+    \frametitle{Conclusion}
     \begin{itemize}

### 2014-03-30 iptables/presentation.tex
rename iptables presentation

--- 2014-03-28 iptables_newline2014/presentation.tex
@@
 \author{Dieter Adriaenssens}
-\date[29Mar2014]{Newline 0x04\\
+\date[Newline 0x04 29Mar2014]{Newline 0x04\\
 Whitespace, Ghent\\
@@
   \begin{frame}
+    \frametitle{Intermezzo : firewall on your PC}
+    Protect your own PC, block all incoming requests :
+    \begin{example}
+      \small{iptables -P INPUT DROP\\
+      iptables -P OUTPUT ACCEPT}\\
+      \pause
+      \small{iptables -A INPUT -i lo -j ACCEPT \# allow local traffic\\
+      iptables -A INPUT -p icmp -j ACCEPT \# allow ping\\
+      iptables -A INPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}
+    \end{example}
+    Add ESTABLISHED, to allow replies of outgoing connections you intiated!
+  \end{frame}
+  \begin{frame}
     \frametitle{Realworld example : setup}
@@
       \item A SMB service is limited to admin and webmaster PCs.
-      \item Some outgoing connections are allowed (dns, dhcp, smtp, ntp).
-      \item Some external websites are reachable.
+      %\item Some outgoing connections are allowed (dns, dhcp, smtp, ntp).
+      %\item Some external websites are reachable.
     \end{itemize}
@@
     \pause
-    The same IP addresses are repeated. If there is a change, several rules have to be updated.\\
+    The same IP addresses are repeated. If there is a change in the IP addresses list, several rules have to be updated.\\
     \pause
@@
     \frametitle{Introducing chains}
-    Firstly, create the chain and add the rules
+    Create the chain and add the rules
     \begin{example}
       \small{iptables -N admin\_IP\\
+      \pause
       iptables -A admin\_IP -s 10.100.2.3 -j ACCEPT\\
@@
     Let's do the same for webmasters AND admins having access to SMB.\\
+    \pause
     Create a chain with webmaster IP addresses :

### 2014-04-18 iptables/presentation.tex
remove leading iptables in config examples

@@
     \begin{example}
-      \small{iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport http -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      iptables -A INPUT -j DROP}
+      \small{-A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport http -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      -A INPUT -j DROP}
     \end{example}
@@
     \begin{example}
-      \small{iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      iptables -A INPUT -j DROP}
+      \small{-A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      -A INPUT -j DROP}
     \end{example}
@@
     \begin{example}
-      \small{iptables -P INPUT DROP\\
-      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      \sout{iptables -A INPUT -j DROP}}
+      \small{-P INPUT DROP\\
+      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      \sout{-A INPUT -j DROP}}
     \end{example}
@@
     \begin{example}
-      \small{iptables -P INPUT ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
+      \small{-P INPUT ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -P INPUT DROP\\
-      iptables -P OUTPUT DROP\\
-      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
+      \small{-P INPUT DROP\\
+      -P OUTPUT DROP\\
+      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -P INPUT DROP\\
-      iptables -P OUTPUT DROP\\
-      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      iptables -A OUTPUT -m tcp -p tcp --sport ssh -j ACCEPT\\
-      iptables -A OUTPUT -m tcp -p tcp --sport 80 -j ACCEPT\\
-      iptables -A OUTPUT -m tcp -p tcp --sport https -j ACCEPT}
+      \small{-P INPUT DROP\\
+      -P OUTPUT DROP\\
+      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      -A OUTPUT -m tcp -p tcp --sport ssh -j ACCEPT\\
+      -A OUTPUT -m tcp -p tcp --sport 80 -j ACCEPT\\
+      -A OUTPUT -m tcp -p tcp --sport https -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -P INPUT DROP\\
-      iptables -P OUTPUT DROP\\
-      iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      iptables -A OUTPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}
+      \small{-P INPUT DROP\\
+      -P OUTPUT DROP\\
+      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      -A OUTPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -P INPUT DROP\\
-      iptables -P OUTPUT ACCEPT}\\
+      \small{-P INPUT DROP\\
+      -P OUTPUT ACCEPT}\\
       \pause
-      \small{iptables -A INPUT -i lo -j ACCEPT \# allow local traffic\\
-      iptables -A INPUT -p icmp -j ACCEPT \# allow ping\\
-      iptables -A INPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}
+      \small{-A INPUT -i lo -j ACCEPT \# allow local traffic\\
+      -A INPUT -p icmp -j ACCEPT \# allow ping\\
+      -A INPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
+      \small{-A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 22 -j ACCEPT\\
-      iptables -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 22 -j ACCEPT\\
-      iptables -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 22 -j ACCEPT\\
-      iptables -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 10000 -j ACCEPT\\
-      iptables -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 10000 -j ACCEPT\\
-      iptables -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 10000 -j ACCEPT}
+      \small{-A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 22 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 22 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 22 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 10000 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 10000 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 10000 -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -N admin\_IP\\
+      \small{-N admin\_IP\\
       \pause
-      iptables -A admin\_IP -s 10.100.2.3 -j ACCEPT\\
-      iptables -A admin\_IP -s 10.100.2.4 -j ACCEPT\\
-      iptables -A admin\_IP -s 10.100.2.7 -j ACCEPT\\
-      iptables -A admin\_IP -j DROP}
+      -A admin\_IP -s 10.100.2.3 -j ACCEPT\\
+      -A admin\_IP -s 10.100.2.4 -j ACCEPT\\
+      -A admin\_IP -s 10.100.2.7 -j ACCEPT\\
+      -A admin\_IP -j DROP}
     \end{example}
@@
     \begin{example}
-      \small{iptables -A INPUT -p tcp -m tcp --dport 22 -j admin\_IP\\
-      iptables -A INPUT -p tcp -m tcp --dport 10000 -j admin\_IP}
+      \small{-A INPUT -p tcp -m tcp --dport 22 -j admin\_IP\\
+      -A INPUT -p tcp -m tcp --dport 10000 -j admin\_IP}
     \end{example}
@@
     \begin{example}
-      \small{iptables -N webmaster\_IP\\
-      iptables -A webmaster\_IP -s 10.100.2.11 -j ACCEPT\\
-      iptables -A webmaster\_IP -s 10.100.2.17 -j ACCEPT\\
-      iptables -A webmaster\_IP -s 10.100.2.34 -j ACCEPT\\
-      iptables -A webmaster\_IP -j DROP}
+      \small{-N webmaster\_IP\\
+      -A webmaster\_IP -s 10.100.2.11 -j ACCEPT\\
+      -A webmaster\_IP -s 10.100.2.17 -j ACCEPT\\
+      -A webmaster\_IP -s 10.100.2.34 -j ACCEPT\\
+      -A webmaster\_IP -j DROP}
     \end{example}
@@
     \begin{example}
-      \small{iptables -A INPUT -p tcp -m tcp --dport 445 -j admin\_IP\\
-      iptables -A INPUT -p tcp -m tcp --dport 445 -j webmaster\_IP\\
-      iptables -A INPUT -p tcp -m tcp --dport 139 -j admin\_IP\\
-      iptables -A INPUT -p tcp -m tcp --dport 139 -j webmaster\_IP}
+      \small{-A INPUT -p tcp -m tcp --dport 445 -j admin\_IP\\
+      -A INPUT -p tcp -m tcp --dport 445 -j webmaster\_IP\\
+      -A INPUT -p tcp -m tcp --dport 139 -j admin\_IP\\
+      -A INPUT -p tcp -m tcp --dport 139 -j webmaster\_IP}
     \end{example}
@@
     \begin{example}
-      \small{iptables -N webmaster\_IP\\
-      iptables -A webmaster\_IP -s 10.100.2.11 -j ACCEPT\\
-      iptables -A webmaster\_IP -s 10.100.2.17 -j ACCEPT\\
-      iptables -A webmaster\_IP -s 10.100.2.34 -j ACCEPT\\
-      \sout{iptables -A webmaster\_IP -j DROP}\\
-      iptables -A webmaster\_IP -j admin\_IP}
+      \small{-N webmaster\_IP\\
+      -A webmaster\_IP -s 10.100.2.11 -j ACCEPT\\
+      -A webmaster\_IP -s 10.100.2.17 -j ACCEPT\\
+      -A webmaster\_IP -s 10.100.2.34 -j ACCEPT\\
+      \sout{-A webmaster\_IP -j DROP}\\
+      -A webmaster\_IP -j admin\_IP}
     \end{example}
@@
     \begin{example}
-      \small{\sout{iptables -A INPUT -p tcp -m tcp --dport 445 -j admin\_IP}\\
-      iptables -A INPUT -p tcp -m tcp --dport 445 -j webmaster\_IP\\
-      \sout{iptables -A INPUT -p tcp -m tcp --dport 139 -j admin\_IP}\\
-      iptables -A INPUT -p tcp -m tcp --dport 139 -j webmaster\_IP}
+      \small{\sout{-A INPUT -p tcp -m tcp --dport 445 -j admin\_IP}\\
+      -A INPUT -p tcp -m tcp --dport 445 -j webmaster\_IP\\
+      \sout{-A INPUT -p tcp -m tcp --dport 139 -j admin\_IP}\\
+      -A INPUT -p tcp -m tcp --dport 139 -j webmaster\_IP}
     \end{example}

### 2014-04-20 iptables/presentation.tex
update event data
add more examples to iptables introduction

@@
 \author{Dieter Adriaenssens}
-\date[Newline 0x04 29Mar2014]{Newline 0x04\\
-Whitespace, Ghent\\
-March 29th, 2014}
+\institute[UGent]{Ghent University, Belgium}
+\date[LinuxTag 8May2014]{LinuxTag - Berlin\\
+May 8th, 2014}
 \subject{iptables}
@@
     \begin{example}
-      \small{-A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport http -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      -A INPUT -j DROP}
+      \small{iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport http -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      iptables -A INPUT -j DROP}
     \end{example}
@@
     \end{itemize}
+    \begin{example}
+      \small{iptables -A \textbf{INPUT} -m tcp -p tcp --dport ssh -j ACCEPT}
+    \end{example}
     You can define your own chains
@@
     \begin{example}
-      \small{iptables -A INPUT -m tcp -p tcp --dport ssh -s 10.0.0.0/8 -j ACCEPT}
+      \small{iptables -A INPUT \textbf{-m tcp -p tcp --dport ssh -s 10.0.0.0/8} -j ACCEPT}
     \end{example}
@@
     \end{itemize}
+    \begin{example}
+      \small{iptables -A INPUT -m tcp -p tcp --dport ssh \textbf{-j ACCEPT}}
+    \end{example}
   \end{frame}
@@
     \begin{example}
-      \small{-A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      -A INPUT -j DROP}
+      \small{iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      iptables -A INPUT -j DROP}
     \end{example}

### 2014-04-20 iptables/presentation.tex
improve sentence
add overview
use Tutorial example as description
add remark about order of the rules

@@
   \titlepage
+  \end{frame}
+  \begin{frame}
+    \frametitle{Overview}
+    \begin{itemize}
+      \item introduction to iptables
+      \item optimizing configuration with an example setup
+    \end{itemize}
   \end{frame}
@@
     \end{example}
-    The last rule matches everything, so if a packet didn't match a previous rule, it will be rejected.
+    The last rule matches everything, so if a packet didn't match a previous rule, it will be rejected.\\
+    \pause
+    \textbf{Remark:} the order of the rules is important
  \end{frame}
@@
   \begin{frame}
-    \frametitle{Realworld example : setup}
+    \frametitle{Tutorial example : setup}
     \begin{itemize}
@@
   \begin{frame}
-    \frametitle{Realworld example : webserver}
+    \frametitle{Tutorial example : webserver}
     Webservice, tcp port 80 (http) and 443 (https) is available for all :
@@
   \begin{frame}
-    \frametitle{Realworld example : sysadmin PCs}
+    \frametitle{Tutorial example : sysadmin PCs}
     Sysadmins are allowed to access SSH (tcp port 22) and a webbased admin tool (tcp port 10000).\\
@@
   \begin{frame}
-    \frametitle{Realworld example : sysadmin PCs}
+    \frametitle{Tutorial example : sysadmin PCs}
     \begin{example}
@@
   \begin{frame}
-    \frametitle{Realworld example : sysadmin PCs}
+    \frametitle{Tutorial example : sysadmin PCs}
     Wouldn't it be convenient to :
@@
   \begin{frame}
-    \frametitle{Realworld example : sysadmin PCs}
+    \frametitle{Tutorial example : sysadmin PCs}
     Add the rules for TCP port 22 and 10000 using the custom chain :
@@
   \begin{frame}
-    \frametitle{Realworld example : webmaster PCs}
+    \frametitle{Tutorial example : webmaster PCs}
     Let's do the same for webmasters AND admins having access to SMB.\\
@@
   \begin{frame}
-    \frametitle{Realworld example : access to SMB}
+    \frametitle{Tutorial example : access to SMB}
     Add the rules for SMB (tcp port 139 and 445) using the custom chains :
@@
     Jump to the sysadmin chain at the end of the webmaster chain.\\
-    \textbf{Warning :} check conditions of chain ending
+    \textbf{Warning :} check conditions of ending chains
   \end{frame}
   \begin{frame}
-    \frametitle{Realworld example : access to SMB}
+    \frametitle{Tutorial example : access to SMB}
     The rules for SMB (tcp port 139 and 445) can be reduced to :

### 2014-04-20 iptables/presentation.tex
remove leading iptables in examples

@@
     \begin{example}
-      \small{iptables -A \textbf{INPUT} -m tcp -p tcp --dport ssh -j ACCEPT}
+      \small{-A \textbf{INPUT} -m tcp -p tcp --dport ssh -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -A INPUT \textbf{-m tcp -p tcp --dport ssh -s 10.0.0.0/8} -j ACCEPT}
+      \small{-A INPUT \textbf{-m tcp -p tcp --dport ssh -s 10.0.0.0/8} -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{iptables -A INPUT -m tcp -p tcp --dport ssh \textbf{-j ACCEPT}}
+      \small{-A INPUT -m tcp -p tcp --dport ssh \textbf{-j ACCEPT}}
     \end{example}

### 2014-04-21 iptables/presentation.tex
use a capital I at the beginning of a sentence
highlight added and removed rules
add 'Example' to title

@@
     %(question to audience : who knows/uses iptables?)
-    iptables is a tool for creating the rulesets for netfilter, a packet filtering framework which was introduced in the linux 2.4 kernel
+    Iptables is a tool for creating the rulesets for netfilter, a packet filtering framework which was introduced in the linux 2.4 kernel
   \end{frame}
@@
     \begin{example}
-      \small{-P INPUT DROP\\
+      \small{{\color{blue}-P INPUT DROP}\\
       -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
@@
       -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      \sout{-A INPUT -j DROP}}
+      \sout{{\color{red}-A INPUT -j DROP}}}
     \end{example}
@@
   \begin{frame}
-    \frametitle{Firewall for your server}
+    \frametitle{Example : firewall for your server}
     Filter both incoming and outgoing traffic :
@@
   \begin{frame}
-    \frametitle{Firewall for your server}
+    \frametitle{Example : firewall for your server}
     Also add rules for outgoing traffic :
@@
       -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      -A OUTPUT -m tcp -p tcp --sport ssh -j ACCEPT\\
+      {\color{blue}-A OUTPUT -m tcp -p tcp --sport ssh -j ACCEPT\\
       -A OUTPUT -m tcp -p tcp --sport 80 -j ACCEPT\\
-      -A OUTPUT -m tcp -p tcp --sport https -j ACCEPT}
+      -A OUTPUT -m tcp -p tcp --sport https -j ACCEPT}}
     \end{example}
@@
       -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      -A OUTPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}
+      {\color{blue}-A OUTPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}}
     \end{example}
@@
       -A webmaster\_IP -s 10.100.2.34 -j ACCEPT\\
-      \sout{-A webmaster\_IP -j DROP}\\
-      -A webmaster\_IP -j admin\_IP}
+      \sout{{\color{red}-A webmaster\_IP -j DROP}}\\
+      {\color{blue}-A webmaster\_IP -j admin\_IP}}
     \end{example}
@@
     \begin{example}
-      \small{\sout{-A INPUT -p tcp -m tcp --dport 445 -j admin\_IP}\\
+      \small{\sout{{\color{red}-A INPUT -p tcp -m tcp --dport 445 -j admin\_IP}}\\
       -A INPUT -p tcp -m tcp --dport 445 -j webmaster\_IP\\
-      \sout{-A INPUT -p tcp -m tcp --dport 139 -j admin\_IP}\\
+      \sout{{\color{red}-A INPUT -p tcp -m tcp --dport 139 -j admin\_IP}}\\
       -A INPUT -p tcp -m tcp --dport 139 -j webmaster\_IP}

### 2014-05-01 iptables/paper.tex
add initial paper for iptables talk at linuxtag

\documentclass[a4paper,12pt]{article}
\usepackage{hyperref}
\usepackage{float}
\usepackage{graphicx}
\usepackage{caption}
\begin{document}
\title{Reducing iptables configuration complexity using chains}
\author{Dieter Adriaenssens\\
Ghent University, Belgium}
\date{LinuxTag - Berlin\\
May 8th, 2014}
\maketitle

\tableofcontents

\section{Abstract}

The goal is to show how to reduce complexity of an iptables configuration, by adding structure to it using chains and as a result making it more maintainable, and faster.

Iptables is the most used local firewall configuring system on linux systems. It is basically a tool for creating the rulesets for netfilter, a packet filtering framework which was introduced in the linux 2.4 kernel. After first explaining the basics about how iptables/netfilter works (rules, targets, the different filters you can set), we get on to the concept of chains, mentioning the default chains (INPUT, OUTPUT, FORWARD) and that custom chains can be created.

Then by using some examples, it is explained how to define different rules to filter traffic and how this can quickly become complex and hard to maintain. By grouping rules with similar parameters (fe. the same protocol and/or port), the rule could filter on the similar part and then target a custom chain, where a set of rules is created that only filter the specific information (like an IP address). The targetted chains can even be combined or reused for several rules. But when doing so one should be beware of the way netfilter behaves when the end of a chain is reached.

The result is a set of rules that looks less complex, because you get a better structured set of rules that point to different chains (with appropriately chosen names). It is more maintainable, because adding a new rule (fe. a new IP address that is allowed to connect to TCP port 22), means just adding a rule that filters on that IP address to the right chain. And it runs faster because only a subset of the filter rules has to be checked.

The testcase used for the examples will be a server setup with some basic services and ports (http(s), ssh, smb and a webbased management tool) and different groups of IP-addresses/ranges that can access those services (general public, sysadmins, webmasters). In the example, several chains will be reused and combined to connect the different chains to the different protocols. Fe. the sysadmins should have access to ssh, smb and the webbased management tool, while the webmasters only need smb.

The audience should have a basic understanding of network concepts like protocols, ports, IP-addresses and packets.

\end{document}

### 2014-05-03 iptables/presentation.tex
add conlusions to overview
Add CC license to iptables presentation, close #1
display contact on bottom of page
use macro to display 2 dashes

@@
 \usetheme{Madrid}
+
+% macro for double dash
+\newcommand{\dd}{{\texttt{-{}-}}}
+
+%% Enable license helpers
+\graphicspath{{../cc_beamer/}}
+\input{../cc_beamer/cc_beamer}
+
 \title[Reduce iptables config complexity]{Reducing iptables configuration complexity using chains}
@@
   \titlepage
+  \vfill
+    \begin{center}
+      \CcGroupBySa{0.83}{0.95ex}\\[2.5ex]
+        {\tiny\CcNote{\CcLongnameBySa}}
+        \vspace*{-2.5ex}
+    \end{center}
   \end{frame}
@@
       \item optimizing configuration with an example setup
+      \item conclusions
     \end{itemize}
@@
     \begin{example}
-      \small{iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport http -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      \small{iptables -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp \dd dport http -j ACCEPT\\
+      iptables -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
       iptables -A INPUT -j DROP}
@@
     \begin{example}
-      \small{-A \textbf{INPUT} -m tcp -p tcp --dport ssh -j ACCEPT}
+      \small{-A \textbf{INPUT} -m tcp -p tcp \dd dport ssh -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{-A INPUT \textbf{-m tcp -p tcp --dport ssh -s 10.0.0.0/8} -j ACCEPT}
+      \small{-A INPUT \textbf{-m tcp -p tcp \dd dport ssh -s 10.0.0.0/8} -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{-A INPUT -m tcp -p tcp --dport ssh \textbf{-j ACCEPT}}
+      \small{-A INPUT -m tcp -p tcp \dd dport ssh \textbf{-j ACCEPT}}
     \end{example}
@@
     \begin{example}
-      \small{-A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      \small{-A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
       -A INPUT -j DROP}
@@
       \small{{\color{blue}-P INPUT DROP}\\
-      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
       \sout{{\color{red}-A INPUT -j DROP}}}
@@
       \small{-P INPUT ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
+      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT}
     \end{example}
@@
       -P OUTPUT DROP\\
-      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
+      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT}
     \end{example}
@@
       -P OUTPUT DROP\\
-      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      {\color{blue}-A OUTPUT -m tcp -p tcp --sport ssh -j ACCEPT\\
-      -A OUTPUT -m tcp -p tcp --sport 80 -j ACCEPT\\
-      -A OUTPUT -m tcp -p tcp --sport https -j ACCEPT}}
+      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
+      {\color{blue}-A OUTPUT -m tcp -p tcp \dd sport ssh -j ACCEPT\\
+      -A OUTPUT -m tcp -p tcp \dd sport 80 -j ACCEPT\\
+      -A OUTPUT -m tcp -p tcp \dd sport https -j ACCEPT}}
     \end{example}
@@
       -P OUTPUT DROP\\
-      -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport https -j ACCEPT\\
-      {\color{blue}-A OUTPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}}
+      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
+      {\color{blue}-A OUTPUT -m state \dd state ESTABLISHED,RELATED -j ACCEPT}}
     \end{example}
@@
       -A INPUT -p icmp -j ACCEPT \# allow ping\\
-      -A INPUT -m state --state ESTABLISHED,RELATED -j ACCEPT}
+      -A INPUT -m state \dd state ESTABLISHED,RELATED -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{-A INPUT -m tcp -p tcp --dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp --dport https -j ACCEPT}
+      \small{-A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
+      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{-A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 22 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 22 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 22 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 10000 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 10000 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 10000 -j ACCEPT}
+      \small{-A INPUT -p tcp -m tcp -s 10.100.2.3 \dd dport 22 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.4 \dd dport 22 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.7 \dd dport 22 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.3 \dd dport 10000 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.4 \dd dport 10000 -j ACCEPT\\
+      -A INPUT -p tcp -m tcp -s 10.100.2.7 \dd dport 10000 -j ACCEPT}
     \end{example}
@@
     \begin{example}
-      \small{-A INPUT -p tcp -m tcp --dport 22 -j admin\_IP\\
-      -A INPUT -p tcp -m tcp --dport 10000 -j admin\_IP}
+      \small{-A INPUT -p tcp -m tcp \dd dport 22 -j admin\_IP\\
+      -A INPUT -p tcp -m tcp \dd dport 10000 -j admin\_IP}
     \end{example}
@@
     \begin{example}
-      \small{-A INPUT -p tcp -m tcp --dport 445 -j admin\_IP\\
-      -A INPUT -p tcp -m tcp --dport 445 -j webmaster\_IP\\
-      -A INPUT -p tcp -m tcp --dport 139 -j admin\_IP\\
-      -A INPUT -p tcp -m tcp --dport 139 -j webmaster\_IP}
+      \small{-A INPUT -p tcp -m tcp \dd dport 445 -j admin\_IP\\
+      -A INPUT -p tcp -m tcp \dd dport 445 -j webmaster\_IP\\
+      -A INPUT -p tcp -m tcp \dd dport 139 -j admin\_IP\\
+      -A INPUT -p tcp -m tcp \dd dport 139 -j webmaster\_IP}
     \end{example}
@@
     \begin{example}
-      \small{\sout{{\color{red}-A INPUT -p tcp -m tcp --dport 445 -j admin\_IP}}\\
-      -A INPUT -p tcp -m tcp --dport 445 -j webmaster\_IP\\
-      \sout{{\color{red}-A INPUT -p tcp -m tcp --dport 139 -j admin\_IP}}\\
-      -A INPUT -p tcp -m tcp --dport 139 -j webmaster\_IP}
+      \small{\sout{{\color{red}-A INPUT -p tcp -m tcp \dd dport 445 -j admin\_IP}}\\
+      -A INPUT -p tcp -m tcp \dd dport 445 -j webmaster\_IP\\
+      \sout{{\color{red}-A INPUT -p tcp -m tcp \dd dport 139 -j admin\_IP}}\\
+      -A INPUT -p tcp -m tcp \dd dport 139 -j webmaster\_IP}
     \end{example}
@@
     Thanks for your attention!\\
-    Questions?\\
+    Questions?
+    \vfill
     Contact

### 2014-05-03 iptables/paper.tex
add a few pagebreaks
add introduction

@@
 
-\tableofcontents
-
+\pagebreak
 \section{Abstract}
@@
 
+\pagebreak
+\tableofcontents
+
+\section{Introduction}
+Setting up a firewall on your *nix box, being it a workstation, laptop, or server, is always a good idea.
+In most cases, you can do with some simple firewall rules, f.e.
+on your laptop, block all incoming requests (except the established connections, i.e.
+the replies on the outgoing requests you made), or on a simple webserver (allow port 80 only).\\
+
+But if you need more complex rules, f.e.
+a server that hosts a website available for the entire internet, but with an ssh and samba service that should only be available for the local subnet, or even some specific IP addresses, it becomes a bit more complex.\\
+And if you want to filter the outgoing traffic as well, your iptables rules get a mess after a while, and when you want to change anything, chances of a mistake or forgetting something are high, which may result in locking yourself out of your box (at least for remote access), or leaving something open that shouldn't.\\
+
+To make your rules more manageable, you can make use of chains in your iptables rules.
+This paper is inspired by an article that uses chains to make iptables more efficient (faster).
+The goal of this paper is to get iptables rules that are easier to read and configure, but it will result in faster handling of packets as well.\\
+\section{Brief introduction to iptables}
+\section{Optimizing configuration with an example setup}
+\section{Conclusion}
+
 \end{document}

### 2014-05-03 iptables/paper.tex
convert to linux format
add subsection about established state

@@
 \section{Brief introduction to iptables}
+\subsection{ESTABLISHED state}
+When using this option, you can filter for established connections.
+If you define it in both the INPUT and OUTPUT rules, you only have to define in the INPUT rules which NEW incoming requests should be allowed, and in the OUTPUT rules which NEW outgoing request are allowed.
+The established connections will be allowed and should not be redefined (making the configuration a lot more readable and maintainable).
+An example allowing only an ssh service without using the ESTABLISHED state would be :
+\begin{verbatim}
+  iptables -A INPUT -p tcp --dport ssh -j ACCEPT
+  iptables -A INPUT -j REJECT
+  iptables -A OUTPUT -p tcp --sport ssh -j ACCEPT
+  iptables -A OUTPUT -j REJECT}
+\end{verbatim}
+
+Basically, every incoming/outgoing connection is dropped, except if the incoming packet has port 22 (ssh) as destination, or if the outgoing packet was sent from port 22 (which is the reply of the ssh server).\\
+
+When using ESTABLISHED state, this will become :
+\begin{verbatim}
+  iptables -A INPUT -p tcp --dport ssh -j ACCEPT
+  iptables -A INPUT -j REJECT
+  iptables -A OUTPUT -m state --state ESTABLISHED -j ACCEPT
+  iptables -A OUTPUT -j REJECT
+\end{verbatim}
+
+Now, every incoming/outgoing connection is dropped, except if the incoming packet has port 22 (ssh) as destination, or if the packet belongs to an established connection.
+Because incoming connections to port 22 are allowed, the firewall will remember a packet coming in, creating a 'connection' for the host/port the packet originates from when the ssh server replies to it.
+So when the reply of the ssh server is sent out, it matches an 'established' connection and will be allowed out.\\
+
+In this example, the benefit of using the connection state is not clear, but when more allowed incoming services are added, they only have to be added on the INPUT chain, but not on the OUTPUT chain, because they are covered by the ESTABLISHED rule.\\
+In the first example (without the ESTABLISHED rule), every allowed incoming connection should be repeated in the OUTPUT chain, matching the packets sent for the outgoing connection, which results in an equal amount of rules on both chains.\\
+If you want to do filtering in both directions (allowing incoming request for listening services and outgoing request for remote services), this can become very messy, and almost unmaintainable without making mistakes.\\
 \section{Optimizing configuration with an example setup}

### 2014-05-03 iptables/paper.tex
add section introducing chains

@@
 The goal of this paper is to get iptables rules that are easier to read and configure, but it will result in faster handling of packets as well.\\
+
 \section{Brief introduction to iptables}
@@
 If you want to do filtering in both directions (allowing incoming request for listening services and outgoing request for remote services), this can become very messy, and almost unmaintainable without making mistakes.\\
+
+\subsection{Introducing chains}
+When two services (on different ports) should be available to a limited but identical list of IP addresses.
+Without using chains, for every combination of port and IP a rule should be created :
+\begin{verbatim}
+  iptables -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 22 -j ACCEPT
+  iptables -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 22 -j ACCEPT
+  iptables -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 22 -j ACCEPT
+
+  iptables -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 10000 -j ACCEPT
+  iptables -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 10000 -j ACCEPT
+  iptables -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 10000 -j ACCEPT
+\end{verbatim}
+
+Resulting in a lot of rules, and when an IP address has to be changed, added or removed, this has to be done for every corresponding rule.\\
+
+When using chains, this can be much easier. Imagine, that you first check if the packet matches the destination port, and if it does, jump to a new chain, where a list of IP addresses is checked. :
+\begin{verbatim}
+  // create new chain admin_IP
+  iptables -N admin_IP
+
+  // add rules to chain admin_IP
+  iptables -A admin_IP -s 10.100.2.3 -j ACCEPT
+  iptables -A admin_IP -s 10.100.2.4 -j ACCEPT
+  iptables -A admin_IP -s 10.100.2.7 -j ACCEPT
+  // drop all packets that are not matched by previous rules
+  iptables -A admin_IP -j DROP
+
+  // filter ports in INPUT chain
+  iptables -A INPUT -p tcp -m tcp --dport 22 -j admin_IP
+  iptables -A INPUT -p tcp -m tcp --dport 10000 -j admin_IP
+\end{verbatim}
+
+As you can see, there is are several benefits of putting the IP addresses in a separate chain :
+\begin{itemize}
+    \item the list of IP addresses in the separate chain can be reused for both ports, so they have to be defined only once.
+    \item adding/changing/removing an IP address is much easier
+    \item there is a better overview of the firewall rules
+\end{itemize}
+
 \section{Optimizing configuration with an example setup}

### 2014-05-03 iptables/paper.tex
fix typo

@@
   iptables -A OUTPUT -p tcp --sport ssh -j ACCEPT
-  iptables -A OUTPUT -j REJECT}
+  iptables -A OUTPUT -j REJECT
 \end{verbatim}

### 2014-05-03 iptables/paper.tex
use short abstract

@@
 
-The goal is to show how to reduce complexity of an iptables configuration, by adding structure to it using chains and as a result making it more maintainable, and faster.
-
-Iptables is the most used local firewall configuring system on linux systems.
-It is basically a tool for creating the rulesets for netfilter, a packet filtering framework which was introduced in the linux 2.4 kernel.
-After first explaining the basics about how iptables/netfilter works (rules, targets, the different filters you can set), we get on to the concept of chains, mentioning the default chains (INPUT, OUTPUT, FORWARD) and that custom chains can be created.
-
-Then by using some examples, it is explained how to define different rules to filter traffic and how this can quickly become complex and hard to maintain.
-By grouping rules with similar parameters (fe.
-the same protocol and/or port), the rule could filter on the similar part and then target a custom chain, where a set of rules is created that only filter the specific information (like an IP address).
-The targetted chains can even be combined or reused for several rules.
-But when doing so one should be beware of the way netfilter behaves when the end of a chain is reached.
-
-The result is a set of rules that looks less complex, because you get a better structured set of rules that point to different chains (with appropriately chosen names).
-It is more maintainable, because adding a new rule (fe.
-a new IP address that is allowed to connect to TCP port 22), means just adding a rule that filters on that IP address to the right chain.
-And it runs faster because only a subset of the filter rules has to be checked.
-
-The testcase used for the examples will be a server setup with some basic services and ports (http(s), ssh, smb and a webbased management tool) and different groups of IP-addresses/ranges that can access those services (general public, sysadmins, webmasters).
-In the example, several chains will be reused and combined to connect the different chains to the different protocols.
-Fe.
-the sysadmins should have access to ssh, smb and the webbased management tool, while the webmasters only need smb.
+After a short introduction of iptables, a local firewall for Linux systems, we go more into detail what chains are, and how these can be used to reduce the complexity of an iptables configuration by grouping similar filtering rules.
+This results in a better structured configuration that is easier to maintain and that runs faster.\\
 

### 2014-05-03 iptables/paper.tex
add configuration

@@
 \section{Optimizing configuration with an example setup}
+\subsection{Actual configuration}
+
+\subsubsection{INPUT chain}
+
+\begin{verbatim}
+  iptables -A INPUT -i lo -j ACCEPT
+  iptables -A INPUT -m state --state ESTABLISHED -j ACCEPT
+  iptables -A INPUT -m state --state RELATED -j ACCEPT
+  iptables -A INPUT -p icmp -j icmp_in
+  iptables -A INPUT -p tcp -m tcp --dport 22 -j admin_IP
+  iptables -A INPUT -p tcp -m tcp --dport 10000 -j admin_IP
+  iptables -A INPUT -p tcp -m tcp --dport 80 -j ACCEPT
+  iptables -A INPUT -p tcp -m tcp --dport 443 -j ACCEPT
+  iptables -A INPUT -p tcp -m tcp --dport 139 -j webmaster_IP
+  iptables -A INPUT -p tcp -m tcp --dport 445 -j webmaster_IP
+  iptables -A INPUT -j DROP
+\end{verbatim}
+
+This is the INPUT chain, allowing :
+\begin{itemize}
+  \item all local traffic (not leaving the physical PC)
+  \item established and related connections
+  \item ICMP packets (ping, etc.) are handled in a seperate chain icmp\_in
+  \item some services
+  \begin{itemize}
+    \item ssh (tcp 22) and webmin (tcp 10000) allowed for admins (admin\_IP chain)
+    \item website (tcp 80 and 443) for everybody
+    \item samba (tcp 139 and 445) for webmasters (and admins, see definition of webmaster\_IP chain)
+  \end{itemize}
+  \item everything else is not allowed (dropped)
+\end{itemize}
+
+\subsubsection{OUTPUT chain}
+
+\begin{verbatim}
+  iptables -A OUTPUT -o lo -j ACCEPT
+  iptables -A OUTPUT -m state --state ESTABLISHED -j ACCEPT
+  iptables -A OUTPUT -m state --state RELATED -j ACCEPT
+  iptables -A OUTPUT -p icmp -m icmp --icmp-type 8 -j ACCEPT
+  iptables -A OUTPUT -p udp -m udp --dport 53 -j ACCEPT
+  iptables -A OUTPUT -p udp -m udp --dport 67 -j ACCEPT
+  iptables -A OUTPUT -p tcp -m tcp --dport 25 -j ACCEPT
+  iptables -A OUTPUT -p udp -m udp --dport 123 -j ACCEPT
+  iptables -A OUTPUT -p tcp -m tcp --dport 80 -j ext_websites
+  iptables -A OUTPUT -j DROP
+\end{verbatim}
+
+The OUTPUT chain, allowing :
+\begin{itemize}
+  \item all local traffic (not leaving the physical PC)
+  \item established and related connections
+  \item ICMP replies
+  \item some remote services (destination IP addresses of hosting servers are not checked)
+  \begin{itemize}
+    \item dns (udp 53) server
+    \item dhcp (udp 67)
+    \item smtp (tcp 25)
+    \item ntp (udp 123)
+  \end{itemize}
+  \item external websites (tcp 80), listed in chain ext\_websites
+  \item everything else is not allowed (dropped)
+\end{itemize}
+
+\subsubsection{icmp in chain}
+
+\begin{verbatim}
+  iptables -N icmp_in
+  iptables -A icmp_in -p icmp -m icmp --icmp-type 8 -j ACCEPT
+  iptables -A icmp_in -p icmp -m icmp --icmp-type 0 -j ACCEPT
+  iptables -A icmp_in -p icmp -m icmp --icmp-type 3 -j ACCEPT
+  iptables -A icmp_in -p icmp -m icmp --icmp-type 4 -j ACCEPT
+  iptables -A icmp_in -p icmp -m icmp --icmp-type 11 -j ACCEPT
+  iptables -A icmp_in -p icmp -m icmp --icmp-type 12 -j ACCEPT
+  iptables -A icmp_in -j DROP
+\end{verbatim}
+
+All allowed incoming ICMP message types.
+
+\subsubsection{admin IP chain}
+
+\begin{verbatim}
+  iptables -N admin_IP
+  iptables -A admin_IP -s 10.100.2.3 -j ACCEPT
+  iptables -A admin_IP -s 10.100.2.4 -j ACCEPT
+  iptables -A admin_IP -s 10.100.2.7 -j ACCEPT
+  iptables -A admin_IP -j DROP
+\end{verbatim}
+
+A list of allowed IP addresses of admin PC's. Everything else is not allowed.
+
+\subsubsection{webmaster IP chain}
+
+\begin{verbatim}
+  iptables -N webmaster_IP
+  iptables -A webmaster_IP -s 10.100.2.11 -j ACCEPT
+  iptables -A webmaster_IP -s 10.100.2.17 -j ACCEPT
+  iptables -A webmaster_IP -s 10.100.2.34 -j ACCEPT
+  iptables -A webmaster_IP -s 10.100.2.50 -j ACCEPT
+  iptables -A webmaster_IP -j admin_IP
+\end{verbatim}
+
+A list of allowed IP addresses of webmaster PC's.
+At the end of the list, it jumps to the admin\_IP chain, chaining both chains.
+
+\subsubsection{external websites chain}
+
+\begin{verbatim}
+  iptables -N ext_websites
+  iptables -A ext_websites -d 212.211.132.250 -j ACCEPT
+  iptables -A ext_websites -d 212.211.132.32 -j ACCEPT
+  iptables -A ext_websites -d 195.20.242.89 -j ACCEPT
+  iptables -A ext_websites -d 130.89.149.225 -j ACCEPT
+  iptables -A ext_websites -d 86.59.118.153 -j ACCEPT
+  iptables -A ext_websites -d 130.89.149.227 -j ACCEPT
+  iptables -A ext_websites -d 128.31.0.51 -j ACCEPT
+  iptables -A ext_websites -d 86.59.118.153 -j ACCEPT
+  iptables -A ext_websites -d 67.228.198.100 -j ACCEPT
+  iptables -A ext_websites -d 140.211.166.6 -j ACCEPT
+  iptables -A ext_websites -d 140.211.166.21 -j ACCEPT
+  iptables -A ext_websites -j LOG
+\end{verbatim}
+
+A list of allowed external websites for updates (mirrors of Debian, webmin and Drupal, in this example).
+All other requests for external websites are logged. This can be useful for monitoring : notification of abuse, or if you forgot to add an allowed website.i
+
 \section{Conclusion}

### 2014-05-03 iptables/paper.tex
add example definition

@@
 \section{Optimizing configuration with an example setup}
+\subsection{Example definition}
+\begin{itemize}
+  \item A web service should be available from all networks (i.e. internet) on port 80 (http) and 443 (https)
+  \item The server can be managed remotely using ssh (port 20) and webmin (port 10000), but only from a limited set of IP addresses (admin PC's).
+  \item The server hosts a samba service (several TCP ports), that should only be available from a limited set of IP addresses (admin + webmaster PC's).
+  \item Outgoing connections will be filtered, but some services should be allowed (dns, dhcp, smtp, ntp) and some external websites should be available to get updates.
+\end{itemize}
+
 \subsection{Actual configuration}

### 2014-05-04 iptables/paper.tex
put abstract on same page as contents
add license info
small improvements

@@
 \usepackage{caption}
+
+%% Enable license helpers
+\graphicspath{{../cc_beamer/}}
+\input{../cc_beamer/cc_beamer}
+
 \begin{document}
@@
 \maketitle
+
+% Creative commons license
+\vfill
+\begin{center}
+  \CcGroupBySa{0.83}{0.95ex}\\[2.5ex]
+    {\footnotesize\CcNote{\CcLongnameBySa}}
+  \vspace*{-2.5ex}
+\end{center}
 
@@
 
+\tableofcontents
 \pagebreak
-\tableofcontents
 
@@
   \begin{itemize}
-    \item ssh (tcp 22) and webmin (tcp 10000) allowed for admins (admin\_IP chain)
-    \item website (tcp 80 and 443) for everybody
-    \item samba (tcp 139 and 445) for webmasters (and admins, see definition of webmaster\_IP chain)
+    \item SSH (tcp 22) and webmin (tcp 10000) allowed for admins (admin\_IP chain)
+    \item website (tcp 80 and 443) for everyone
+    \item SMB (tcp 139 and 445) for webmasters (and admins, see definition of webmaster\_IP chain)
   \end{itemize}
@@
   \item established and related connections
-  \item ICMP replies
-  \item some remote services (destination IP addresses of hosting servers are not checked)
+  \item ICMP replies (fe. ping)
+  \item some remote services :
   \begin{itemize}
-    \item dns (udp 53) server
-    \item dhcp (udp 67)
-    \item smtp (tcp 25)
-    \item ntp (udp 123)
+    \item DNS (udp 53) : name service
+    \item DHCP (udp 67) : dynamic IP address service
+    \item SMTP (tcp 25) : mail service
+    \item NTP (udp 123) : time service
   \end{itemize}
@@
 A list of allowed external websites for updates (mirrors of Debian, webmin and Drupal, in this example).
-All other requests for external websites are logged. This can be useful for monitoring : notification of abuse, or if you forgot to add an allowed website.i
+All other requests for external websites are logged. This can be useful for monitoring : notification of abuse, or if you forgot to add an allowed website.
 

### 2014-05-04 iptables/paper.tex
add conclusion

@@
 \section{Conclusion}
+\begin{itemize}
+  \item When checking for the ESTABLISHED state, the number of rules can be reduced
+  \item Using chains makes your rules better structured which makes them easier to read and better maintainable
+  \item Chains can be reused for several rules
+  \item Chains can be chained together
+  \item Using chains makes checking rules faster, because it only jumps to a chain when a rule matches. Not all rules have to be checked.
+\end{itemize}
 

### 2014-05-04 iptables/paper.tex
small changes

@@
 But if you need more complex rules, f.e.
-a server that hosts a website available for the entire internet, but with an ssh and samba service that should only be available for the local subnet, or even some specific IP addresses, it becomes a bit more complex.\\
+a server that hosts a website available for the entire internet, but with an SSH and samba service that should only be available for the local subnet, or even some specific IP addresses, it becomes a bit more complex.\\
 And if you want to filter the outgoing traffic as well, your iptables rules get a mess after a while, and when you want to change anything, chances of a mistake or forgetting something are high, which may result in locking yourself out of your box (at least for remote access), or leaving something open that shouldn't.\\
@@
 The established connections will be allowed and should not be redefined (making the configuration a lot more readable and maintainable).
-An example allowing only an ssh service without using the ESTABLISHED state would be :
+An example allowing only an SSH service without using the ESTABLISHED state would be :
 \begin{verbatim}
@@
 
-Basically, every incoming/outgoing connection is dropped, except if the incoming packet has port 22 (ssh) as destination, or if the outgoing packet was sent from port 22 (which is the reply of the ssh server).\\
+Basically, every incoming/outgoing connection is dropped, except if the incoming packet has port 22 (SSH) as destination, or if the outgoing packet was sent from port 22 (which is the reply of the SSH server).\\
 
@@
 
-Now, every incoming/outgoing connection is dropped, except if the incoming packet has port 22 (ssh) as destination, or if the packet belongs to an established connection.
-Because incoming connections to port 22 are allowed, the firewall will remember a packet coming in, creating a 'connection' for the host/port the packet originates from when the ssh server replies to it.
-So when the reply of the ssh server is sent out, it matches an 'established' connection and will be allowed out.\\
+Now, every incoming/outgoing connection is dropped, except if the incoming packet has port 22 (SSH) as destination, or if the packet belongs to an established connection.
+Because incoming connections to port 22 are allowed, the firewall will remember a packet coming in, creating a 'connection' for the host/port the packet originates from when the SSH server replies to it.
+So when the reply of the SSH server is sent out, it matches an 'established' connection and will be allowed out.\\
 
@@
   \item A web service should be available from all networks (i.e. internet) on port 80 (http) and 443 (https)
-  \item The server can be managed remotely using ssh (port 20) and webmin (port 10000), but only from a limited set of IP addresses (admin PC's).
+  \item The server can be managed remotely using SSH (port 22) and a webbased admin tool (port 10000), but only from a limited set of IP addresses (admin PC's).
   \item The server hosts a samba service (several TCP ports), that should only be available from a limited set of IP addresses (admin + webmaster PC's).
-  \item Outgoing connections will be filtered, but some services should be allowed (dns, dhcp, smtp, ntp) and some external websites should be available to get updates.
+  \item Outgoing connections will be filtered, but some services should be allowed (DNS, DHCP, SMTP, NTP) and some external websites should be available to get updates.
 \end{itemize}

### 2014-05-04 iptables/paper.tex
improve conclusion
add brief introduction to iptables and chains section

@@
 \section{Brief introduction to iptables}
+  Iptables is a tool for creating the rulesets for netfilter, a packet filtering framework which was introduced in the linux 2.4 kernel.\\
+
+  Netfilter is rule based. When an IP packet arrives, it is checked against a set of rules, fe. :
+  \begin{verbatim}
+    iptables -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT
+    iptables -A INPUT -m tcp -p tcp \dd dport http -j ACCEPT
+    iptables -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT
+    iptables -A INPUT -j DROP
+  \end{verbatim}
+
+  Every rule consists of several parts :
+  \begin{itemize}
+    \item the chain the rule belongs to
+    \item parameters filtering for a type of packet
+    \item the target : action to be taken when a packet matches
+  \end{itemize}
+
+  \subsection{Chains}
+    There are several predefined chains :
+    \begin{itemize}
+      \item INPUT
+      \item OUTPUT
+      \item FORWARD
+      \item PREROUTING
+      \item POSTROUTING
+    \end{itemize}
+    Defining custom chains is also possible.\\
+
+    In this example, the rule is added to the INPUT chain :
+    \begin{verbatim}
+iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT
+    \end{verbatim}
+
 \subsection{ESTABLISHED state}
@@
 \begin{itemize}
-  \item When checking for the ESTABLISHED state, the number of rules can be reduced
+  \item When checking for the ESTABLISHED state, the number of rules can be reduced, because only the initiating packet needs to be checked
   \item Using chains makes your rules better structured which makes them easier to read and better maintainable
@@
   \item Chains can be chained together
-  \item Using chains makes checking rules faster, because it only jumps to a chain when a rule matches. Not all rules have to be checked.
+  \item Using chains makes checking rules faster, because it only jumps to a chain when a rule matches. Not all rules have to be checked
 \end{itemize}

### 2014-05-04 iptables/paper.tex
add filters and target subsection placeholders
use default and custom chains

@@
 
-  \subsection{Chains}
+  \subsection{Default chains}
     There are several predefined chains :
@@
     \end{verbatim}
+
+  \subsection{Filters}
+  \subsection{Targets}
 
@@
 
-\subsection{Introducing chains}
+\subsection{Introducing custom chains}
 When two services (on different ports) should be available to a limited but identical list of IP addresses.
@@
 
-When using chains, this can be much easier. Imagine, that you first check if the packet matches the destination port, and if it does, jump to a new chain, where a list of IP addresses is checked. :
+When using custom chains, this can be much easier.
+Imagine, that you first check if the packet matches the destination port, and if it does, jump to a new chain, where a list of IP addresses is checked.
+:
 \begin{verbatim}
-  // create new chain admin_IP
+  // create new custom chain admin_IP
   iptables -N admin_IP
 
-  // add rules to chain admin_IP
+  // add rules to custom chain admin_IP
   iptables -A admin_IP -s 10.100.2.3 -j ACCEPT

### 2014-05-04 iptables/paper.tex
add section on filters
add targets section

@@
   \subsection{Filters}
+
+    Filter on packet properties :
+    \begin{itemize}
+      \item protocol (tcp, udp, icmp, \ldots)
+      \item destination/source port
+      \item destination/source IP address
+      \item in/outgoing interface (eth0, \ldots)
+      \item \ldots
+    \end{itemize}
+
+    In this example, the filter matches any TCP packet with destination port 22 (SSH) originating from the 10.0.0.0/8 network :
+    \begin{verbatim}
+iptables -A INPUT -m tcp -p tcp --dport ssh -s 10.0.0.0/8 -j ACCEPT
+    \end{verbatim}
+
   \subsection{Targets}
+
+    What happens if a packet matches a rule :
+    \begin{itemize}
+      \item ACCEPT
+      \item DROP
+      \item QUEUE $\rightarrow$ userspace
+      \item RETURN $\rightarrow$ leave current chain
+      \item LOG
+      \item jump to a custom chain
+    \end{itemize}
+
+    In this example, the packet is accepted if it matches the rule :
+    \begin{verbatim}
+iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT
+    \end{verbatim}
 

### 2014-05-04 iptables/paper.tex
add section about rule order

@@
     \end{verbatim}
+
+    \subsection{Putting it all together}
+    All rules are checked one for one, and if one matches the target is executed, in this case, the packet is accepted :
+    \begin{verbatim}
+iptables -A INPUT -m tcp -p tcp --dport ssh -j ACCEPT
+iptables -A INPUT -m tcp -p tcp --dport 80 -j ACCEPT
+iptables -A INPUT -m tcp -p tcp --dport https -j ACCEPT
+iptables -A INPUT -j DROP
+    \end{verbatim}
+    The last rule matches everything, so if a packet didn't match a previous rule, it will be rejected.\\
+    \textbf{Remark:} the order of the rules is important.
 

### 2014-05-04 iptables/paper.tex
reorganise some sections
rename some titles
refer to detailed iptables configuration tutorial

@@
 
+\subsection{More about iptables}
+This was just a brief introduction to iptables/netfilter.
+If you want to know more about iptables configuration, you can have a look at this excelent and detailed tutorial : http://www.frozentux.net/iptables-tutorial/iptables-tutorial.html
+\newpage
+
+\section{Optimizing configuration}
+\subsection{Example setup definition}
+\begin{itemize}
+  \item A web service should be available from all networks (i.e. internet) on port 80 (http) and 443 (https)
+  \item The server can be managed remotely using SSH (port 22) and a webbased admin tool (port 10000), but only from a limited set of IP addresses (admin PC's).
+  \item The server hosts a samba service (several TCP ports), that should only be available from a limited set of IP addresses (admin + webmaster PC's).
+  \item Outgoing connections will be filtered, but some services should be allowed (DNS, DHCP, SMTP, NTP) and some external websites should be available to get updates.
+\end{itemize}
+
   \subsection{ESTABLISHED state}
@@
 
-\section{Optimizing configuration with an example setup}
-\subsection{Example definition}
-\begin{itemize}
-  \item A web service should be available from all networks (i.e. internet) on port 80 (http) and 443 (https)
-  \item The server can be managed remotely using SSH (port 22) and a webbased admin tool (port 10000), but only from a limited set of IP addresses (admin PC's).
-  \item The server hosts a samba service (several TCP ports), that should only be available from a limited set of IP addresses (admin + webmaster PC's).
-  \item Outgoing connections will be filtered, but some services should be allowed (DNS, DHCP, SMTP, NTP) and some external websites should be available to get updates.
-\end{itemize}
-
 \subsection{Actual configuration}
@@
 
-\subsubsection{icmp in chain}
+\subsubsection{ICMP chain}
 
@@
 
-\subsubsection{admin IP chain}
+\subsubsection{admin chain}
 
@@
 
-\subsubsection{webmaster IP chain}
+\subsubsection{webmaster chain}
 
@@
 
-\subsubsection{external websites chain}
+\subsubsection{External websites chain}
 

### 2014-05-04 iptables/paper.tex
add section about the author

@@
 
+\section{About the author}
+Dieter Adriaenssens currently works at Ghent University and is a sysadmin since 2005.
+He is an active Open Source community member, who contributed to phpMyAdmin and other projects, regularly attending and speaking at conferences, like FOSDEM.\\
+Recently, Dieter started exploring the world of Android application development, which resulted in a first release of a navigation app a few months ago.\\
+He lives in Ghent, Belgium and enjoys rock climbing.
+\begin{itemize}
+  \item Blog : \href{http://ruleant.blogspot.com/}{http://ruleant.blogspot.com}
+  \item Twitter : \href{https://twitter.com/dcadriaenssens}{@dcadriaenssens}
+\end{itemize}
+
 \end{document}

### 2014-05-04 iptables/paper.tex
make link clickable

@@
 This was just a brief introduction to iptables/netfilter.
-If you want to know more about iptables configuration, you can have a look at this excelent and detailed tutorial : http://www.frozentux.net/iptables-tutorial/iptables-tutorial.html
+If you want to know more about iptables configuration, you can have a look at this excelent and detailed tutorial : \href{http://www.frozentux.net/iptables-tutorial/iptables-tutorial.html}{http://www.frozentux.net/iptables-tutorial/iptables-tutorial.html}
 \newpage

### 2014-05-04 iptables/paper.tex
add author name of iptables tutorial

--- 2014-05-04 iptables/paper.tex
@@
 This was just a brief introduction to iptables/netfilter.
-If you want to know more about iptables configuration, you can have a look at this excelent and detailed tutorial : http://www.frozentux.net/iptables-tutorial/iptables-tutorial.html
+If you want to know more about iptables configuration, you can have a look at this excellent and detailed tutorial by Oskar Andreasson : \href{http://www.frozentux.net/iptables-tutorial/iptables-tutorial.html}{http://www.frozentux.net/iptables-tutorial/iptables-tutorial.html}
 \newpage

### 2014-05-07 iptables/presentation.tex
add 'brief' to introduction
use capitals
add twitter account to first slide
fix cc position
use term 'use 'case'
fix grammar

@@
 \author{Dieter Adriaenssens}
-\institute[UGent]{Ghent University, Belgium}
+\institute[UGent]{@dcadriaenssens - Ghent University, Belgium}
 \date[LinuxTag 8May2014]{LinuxTag - Berlin\\
@@
     \begin{itemize}
-      \item introduction to iptables
-      \item optimizing configuration with an example setup
-      \item conclusions
+      \item Brief introduction to iptables
+      \item Optimizing configuration with a tutorial use case
+      \item Conclusion
     \end{itemize}
@@
     \frametitle{Putting it all together}
-    All rules are checked one for one, and if one matches the target is executed :
+    All rules are checked one by one, and if one matches the target is executed :
     \begin{example}
@@
   \begin{frame}
-    \frametitle{Tutorial example : setup}
+    \frametitle{Tutorial use case : setup}
     \begin{itemize}
@@
   \begin{frame}
-    \frametitle{Tutorial example : webserver}
+    \frametitle{Tutorial use case : webserver}
     Webservice, tcp port 80 (http) and 443 (https) is available for all :
@@
   \begin{frame}
-    \frametitle{Tutorial example : sysadmin PCs}
+    \frametitle{Tutorial use case : sysadmin PCs}
     Sysadmins are allowed to access SSH (tcp port 22) and a webbased admin tool (tcp port 10000).\\
@@
   \begin{frame}
-    \frametitle{Tutorial example : sysadmin PCs}
+    \frametitle{Tutorial use case : sysadmin PCs}
     \begin{example}
@@
   \begin{frame}
-    \frametitle{Tutorial example : sysadmin PCs}
+    \frametitle{Tutorial use case : sysadmin PCs}
     Wouldn't it be convenient to :
@@
   \begin{frame}
-    \frametitle{Tutorial example : sysadmin PCs}
+    \frametitle{Tutorial use case : sysadmin PCs}
     Add the rules for TCP port 22 and 10000 using the custom chain :
@@
   \begin{frame}
-    \frametitle{Tutorial example : webmaster PCs}
+    \frametitle{Tutorial use case : webmaster PCs}
     Let's do the same for webmasters AND admins having access to SMB.\\
@@
   \begin{frame}
-    \frametitle{Tutorial example : access to SMB}
+    \frametitle{Tutorial use case : access to SMB}
     Add the rules for SMB (tcp port 139 and 445) using the custom chains :
@@
   \begin{frame}
-    \frametitle{Tutorial example : access to SMB}
+    \frametitle{Tutorial use case : access to SMB}
     The rules for SMB (tcp port 139 and 445) can be reduced to :

### 2014-05-07 iptables/presentation.tex
use passive tense
add comma to improve readability
also mention other icmp types

@@
     \end{example}
-    You can define your own chains
+    Custom chains can be defined
   \end{frame}
@@
     \frametitle{Putting it all together}
-    All rules are checked one by one, and if one matches the target is executed :
+    All rules are checked one by one, and if one matches, the target is executed :
     \begin{example}
@@
     \frametitle{Default target DROP}
-    But you can define default target for the predefined chains:
+    A default target can be set for the predefined chains:
     \begin{example}
@@
       \small{-A INPUT -i lo -j ACCEPT \# allow local traffic\\
-      -A INPUT -p icmp -j ACCEPT \# allow ping\\
+      -A INPUT -p icmp -j ACCEPT \# allow ping, etc\\
       -A INPUT -m state \dd state ESTABLISHED,RELATED -j ACCEPT}

### 2014-05-07 iptables/presentation.tex
use match, not filter, to better reflect iptables documentation
add slide on IPv6

@@
   \begin{frame}
-    \frametitle{Filters}
+    \frametitle{Match}
     Filter on packet parameters :
@@
   \end{frame}
+  \begin{frame}
+    \frametitle{IPv6}
+    Configuration is the same for IPv4, use ip6tables :
+    \begin{example}
+      \small{ip6tables -P INPUT DROP\\
+      ip6tables -P OUTPUT ACCEPT\\
+      ip6tables -A INPUT -i lo -j ACCEPT \# allow local traffic\\
+      ip6tables -A INPUT -p icmp -j ACCEPT \# allow ping, etc\\
+      ip6tables -A INPUT -m state \dd state ESTABLISHED,RELATED -j ACCEPT}
+    \end{example}
+    Even if IPv4 rules are set, IPv6 has to be configured seperately, otherwise the defaults apply.
+  \end{frame}
 \end{document}

### 2014-05-08 iptables/presentation.tex
mention IPv6 day
create seperate page for IPv6 only day
split lines

--- 2014-05-07 iptables/presentation.tex
@@
   \end{frame}
+  \begin{frame}
+    \frametitle{IPv6}
+    Configuration is the same for IPv4, use ip6tables :
+    \begin{example}
+      \small{ip6tables -P INPUT DROP\\
+      ip6tables -P OUTPUT ACCEPT\\
+      ip6tables -A INPUT -i lo -j ACCEPT \# allow local traffic\\
+      ip6tables -A INPUT -p icmp -j ACCEPT \# allow ping, etc\\
+      ip6tables -A INPUT -m state \dd state ESTABLISHED -j ACCEPT\\
+      ip6tables -A INPUT -m state \dd state RELATED -j ACCEPT}
+    \end{example}
+    Even if IPv4 rules are set, IPv6 has to be configured seperately, otherwise the defaults apply.
+  \end{frame}
+  \begin{frame}
+    \frametitle{IPv6 only day}
+    \begin{columns}
+      \begin{column}{5cm}
+        \includegraphics[scale=.3]{pictures/World_IPv6_launch_badge_512.png}
+      \end{column}
+      \begin{column}{6cm}
+        \textbf{June 6th 2014}\\
+
+        Disable IPv4 for one day!
+      \end{column}
+    \end{columns}
+  \end{frame}
 \end{document}

### 2014-05-08 iptables/presentation.tex
highlight ip6tables
use term 'custom chains' everywhere
mention better performance

@@
     \pause
-    \textbf{This is possible with chains!}
+    \textbf{This is possible with custom chains!}
   \end{frame}
   \begin{frame}
-    \frametitle{Introducing chains}
-    Create the chain and add the rules
+    \frametitle{Introducing custom chains}
+    Create the custom chain and add the rules
     \begin{example}
@@
     \begin{itemize}
-      \item the list of IP addresses in the custom chain is reused for both ports, so they have to be defined only once.
+      \item the list of IP addresses in the custom chain is reused for both ports, so they have to be defined only once
       \item adding/changing/removing an IP address is much easier
-      \item a better overview of the firewall rules.
+      \item a better overview of the firewall rules
+      \item better performance : the rules in the custom chain are only checked if that chain is accessed
     \end{itemize}
@@
     \frametitle{IPv6}
-    Configuration is the same for IPv4, use ip6tables :
+    Configuration is the same for IPv4, use \textit{ip6tables} :
     \begin{example}

### 2014-05-09 iptables/presentation.tex
mention reason for closing custom chain
add pages about using default target for custom chains
mention netfilter with iptables
Revert "create seperate page for IPv6 only day"
This reverts commit bfb9d3acdd39dc891688a60c0649f0c63aca73cc.

@@
     \begin{itemize}
-      \item Brief introduction to iptables
+      \item Brief introduction to iptables/netfilter
       \item Optimizing configuration with a tutorial use case
@@
   \begin{frame}
-    \frametitle{What is iptables?}
+    \frametitle{What is iptables/netfilter?}
     %(question to audience : who knows/uses iptables?)
@@
   \begin{frame}
+    \frametitle{Why you should set a default target}
+    Avoid jumping back to the calling chain :
+    \begin{itemize}
+      \item performance issue
+      \item unexpected behaviour, some other rules might match
+    \end{itemize}
+ \end{frame}
+  \begin{frame}
     \frametitle{Tutorial use case : sysadmin PCs}
@@
     \end{example}
-    Even if IPv4 rules are set, IPv6 has to be configured seperately, otherwise the defaults apply.
-  \end{frame}
-  \begin{frame}
-    \frametitle{IPv6 only day}
-    \begin{columns}
-      \begin{column}{5cm}
-        \includegraphics[scale=.3]{pictures/World_IPv6_launch_badge_512.png}
-      \end{column}
-      \begin{column}{6cm}
-        \textbf{June 6th 2014}\\
-
-        Disable IPv4 for one day!
-      \end{column}
-    \end{columns}
+    Even if IPv4 rules are set, IPv6 has to be configured seperately, otherwise the defaults apply.\\
+    \pause
+    \textbf{IPv6 day (June 6th 2014)} $\rightarrow$ disable IPv4 for one day
   \end{frame}

### 2015-03-22 buildtimetrend/presentation.tex
start Buildtime Trend presentation

--- 2014-05-09 iptables/presentation.tex
@@
 
-\title[Reduce iptables config complexity]{Reducing iptables configuration complexity using chains}
+\title[Buildtime Trend]{Buildtime Trend}
 \author{Dieter Adriaenssens}
-\institute[UGent]{@dcadriaenssens - Ghent University, Belgium}
-\date[LinuxTag 8May2014]{LinuxTag - Berlin\\
-May 8th, 2014}
-\subject{iptables}
+\institute[Buildtime Trend]{@dcadriaenssens - Buildtime Trend}
+\date[NewLine 5Apr2015]{NewLine - Ghent\\
+April 5th, 2015}
+\subject{Buildtime Trend}
 \begin{document}
@@
     \begin{center}
-      \CcGroupBySa{0.83}{0.95ex}\\[2.5ex]
-        {\tiny\CcNote{\CcLongnameBySa}}
+      \CcGroupByNcSa{0.83}{0.95ex}\\[2.5ex]
+        {\tiny\CcNote{\CcLongnameByNcSa}}
         \vspace*{-2.5ex}
@@
     Iptables is a tool for creating the rulesets for netfilter, a packet filtering framework which was introduced in the linux 2.4 kernel
-  \end{frame}
-  \begin{frame}
-    \frametitle{Rules}
-    When an IP packet comes in, it is checked against a set of rules.
-    \begin{example}
-      \small{iptables -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp \dd dport http -j ACCEPT\\
-      iptables -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
-      iptables -A INPUT -j DROP}
-    \end{example}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Chains}
-    Predefined chains :
-    \begin{itemize}
-      \item INPUT
-      \item OUTPUT
-      \item FORWARD
-      \item PREROUTING
-      \item POSTROUTING
-    \end{itemize}
-    \begin{example}
-      \small{-A \textbf{INPUT} -m tcp -p tcp \dd dport ssh -j ACCEPT}
-    \end{example}
-    Custom chains can be defined
-  \end{frame}
-  \begin{frame}
-    \frametitle{Match}
-    Filter on packet parameters :
-    \begin{itemize}
-      \item protocol (tcp, udp, icmp, \ldots)
-      \item destination/source port
-      \item destination/source IP address
-      \item in/outgoing interface (eth0, \ldots)
-      \item \ldots
-    \end{itemize}
-    \begin{example}
-      \small{-A INPUT \textbf{-m tcp -p tcp \dd dport ssh -s 10.0.0.0/8} -j ACCEPT}
-    \end{example}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Targets}
-    What to do if a packet matches a rule :
-    \begin{itemize}
-      \item ACCEPT
-      \item DROP
-      \item QUEUE $\rightarrow$ userspace
-      \item RETURN $\rightarrow$ leave current chain
-      \item LOG
-      \item jump to a custom chain
-    \end{itemize}
-    \begin{example}
-      \small{-A INPUT -m tcp -p tcp \dd dport ssh \textbf{-j ACCEPT}}
-    \end{example}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Putting it all together}
-    All rules are checked one by one, and if one matches, the target is executed :
-    \begin{example}
-      \small{-A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
-      -A INPUT -j DROP}
-    \end{example}
-    The last rule matches everything, so if a packet didn't match a previous rule, it will be rejected.\\
-    \pause
-    \textbf{Remark:} the order of the rules is important
- \end{frame}
-  \begin{frame}
-    \frametitle{Default target DROP}
-    A default target can be set for the predefined chains:
-    \begin{example}
-      \small{{\color{blue}-P INPUT DROP}\\
-      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
-      \sout{{\color{red}-A INPUT -j DROP}}}
-    \end{example}
-    If none of the rules match, the default target is executed.\\
-    In this example, a packet is dropped.
-  \end{frame}
-  \begin{frame}
-    \frametitle{Default target ACCEPT}
-    \textbf{Exercise:}\\
-    What happens if the default target is ACCEPT?
-    \begin{example}
-      \small{-P INPUT ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT}
-    \end{example}
-    \pause
-    All packets are accepted, even if no rule matches them.\\
-    \pause
-    \textbf{Always define a default target or use an all-matching rule at the end!}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Example : firewall for your server}
-    Filter both incoming and outgoing traffic :
-    \begin{example}
-      \small{-P INPUT DROP\\
-      -P OUTPUT DROP\\
-      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT}
-    \end{example}
-    \pause
-    \textbf{Problem:} incoming packets are accepted, but replies are dropped!
-  \end{frame}
-  \begin{frame}
-    \frametitle{Example : firewall for your server}
-    Also add rules for outgoing traffic :
-    \begin{example}
-      \small{-P INPUT DROP\\
-      -P OUTPUT DROP\\
-      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
-      {\color{blue}-A OUTPUT -m tcp -p tcp \dd sport ssh -j ACCEPT\\
-      -A OUTPUT -m tcp -p tcp \dd sport 80 -j ACCEPT\\
-      -A OUTPUT -m tcp -p tcp \dd sport https -j ACCEPT}}
-    \end{example}
-    \pause
-    This can get complicated if more rules are added.
-  \end{frame}
-  \begin{frame}
-    \frametitle{Established state}
-    \textbf{Solution:} check for established state
-    \begin{example}
-      \small{-P INPUT DROP\\
-      -P OUTPUT DROP\\
-      -A INPUT -m tcp -p tcp \dd dport ssh -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT\\
-      {\color{blue}-A OUTPUT -m state \dd state ESTABLISHED,RELATED -j ACCEPT}}
-    \end{example}
-    This makes the rules more manageable, especially when output rules are also defined.
-  \end{frame}
-  \begin{frame}
-    \frametitle{Intermezzo : firewall on your PC}
-    Protect your own PC, block all incoming requests :
-    \begin{example}
-      \small{-P INPUT DROP\\
-      -P OUTPUT ACCEPT}\\
-      \pause
-      \small{-A INPUT -i lo -j ACCEPT \# allow local traffic\\
-      -A INPUT -p icmp -j ACCEPT \# allow ping, etc\\
-      -A INPUT -m state \dd state ESTABLISHED,RELATED -j ACCEPT}
-    \end{example}
-    Add ESTABLISHED, to allow replies of outgoing connections you intiated!
-  \end{frame}
-  \begin{frame}
-    \frametitle{Tutorial use case : setup}
-    \begin{itemize}
-      \item Tcp port 80 (http) and 443 (https) is available for all.
-      \item SSH (port 22) and a webbased admin tool (port 10000), is limited to admin PCs.
-      \item A SMB service is limited to admin and webmaster PCs.
-      %\item Some outgoing connections are allowed (dns, dhcp, smtp, ntp).
-      %\item Some external websites are reachable.
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Tutorial use case : webserver}
-    Webservice, tcp port 80 (http) and 443 (https) is available for all :
-    \begin{example}
-      \small{-A INPUT -m tcp -p tcp \dd dport 80 -j ACCEPT\\
-      -A INPUT -m tcp -p tcp \dd dport https -j ACCEPT}
-    \end{example}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Tutorial use case : sysadmin PCs}
-    Sysadmins are allowed to access SSH (tcp port 22) and a webbased admin tool (tcp port 10000).\\
-    The IP addresses of their PCs :
-    \begin{itemize}
-	\item 10.100.2.3
-	\item 10.100.2.4
-	\item 10.100.2.7
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Tutorial use case : sysadmin PCs}
-    \begin{example}
-      \small{-A INPUT -p tcp -m tcp -s 10.100.2.3 \dd dport 22 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.4 \dd dport 22 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.7 \dd dport 22 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.3 \dd dport 10000 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.4 \dd dport 10000 -j ACCEPT\\
-      -A INPUT -p tcp -m tcp -s 10.100.2.7 \dd dport 10000 -j ACCEPT}
-    \end{example}
-    \pause
-    The same IP addresses are repeated. If there is a change in the IP addresses list, several rules have to be updated.\\
-    \pause
-    \textbf{This can be done more efficient!}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Tutorial use case : sysadmin PCs}
-    Wouldn't it be convenient to :
-    \begin{itemize}
-      \item make a list of IP addresses?
-      \item check this list when a packet matches a rule (fe. TCP port 22)?
-      \item reuse this list when another rule matches a packet?
-    \end{itemize}
-    \pause
-    \textbf{This is possible with custom chains!}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Introducing custom chains}
-    Create the custom chain and add the rules
-    \begin{example}
-      \small{-N admin\_IP\\
-      \pause
-      -A admin\_IP -s 10.100.2.3 -j ACCEPT\\
-      -A admin\_IP -s 10.100.2.4 -j ACCEPT\\
-      -A admin\_IP -s 10.100.2.7 -j ACCEPT\\
-      -A admin\_IP -j DROP}
-    \end{example}
-    \pause
-    \begin{itemize}[<+->]
-      \item Syntax of adding rules to a custom chain is similar to adding to default targets.
-      \item Custom chains don't have a default target, so set a target for all packets that are not matched.
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Why you should set a default target}
-    Avoid jumping back to the calling chain :
-    \begin{itemize}
-      \item performance issue
-      \item unexpected behaviour, some other rules might match
-    \end{itemize}
- \end{frame}
-  \begin{frame}
-    \frametitle{Tutorial use case : sysadmin PCs}
-    Add the rules for TCP port 22 and 10000 using the custom chain :
-    \begin{example}
-      \small{-A INPUT -p tcp -m tcp \dd dport 22 -j admin\_IP\\
-      -A INPUT -p tcp -m tcp \dd dport 10000 -j admin\_IP}
-    \end{example}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Benefits}
-    \begin{itemize}
-      \item the list of IP addresses in the custom chain is reused for both ports, so they have to be defined only once
-      \item adding/changing/removing an IP address is much easier
-      \item a better overview of the firewall rules
-      \item better performance : the rules in the custom chain are only checked if that chain is accessed
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Tutorial use case : webmaster PCs}
-    Let's do the same for webmasters AND admins having access to SMB.\\
-    \pause
-    Create a chain with webmaster IP addresses :
-    \begin{example}
-      \small{-N webmaster\_IP\\
-      -A webmaster\_IP -s 10.100.2.11 -j ACCEPT\\
-      -A webmaster\_IP -s 10.100.2.17 -j ACCEPT\\
-      -A webmaster\_IP -s 10.100.2.34 -j ACCEPT\\
-      -A webmaster\_IP -j DROP}
-    \end{example}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Tutorial use case : access to SMB}
-    Add the rules for SMB (tcp port 139 and 445) using the custom chains :
-    \begin{example}
-      \small{-A INPUT -p tcp -m tcp \dd dport 445 -j admin\_IP\\
-      -A INPUT -p tcp -m tcp \dd dport 445 -j webmaster\_IP\\
-      -A INPUT -p tcp -m tcp \dd dport 139 -j admin\_IP\\
-      -A INPUT -p tcp -m tcp \dd dport 139 -j webmaster\_IP}
-    \end{example}
-  \pause
-  \textbf{This can be optimized!}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Chaining chains}
-    Include the sysadmin chain in the webmaster chain:
-    \begin{example}
-      \small{-N webmaster\_IP\\
-      -A webmaster\_IP -s 10.100.2.11 -j ACCEPT\\
-      -A webmaster\_IP -s 10.100.2.17 -j ACCEPT\\
-      -A webmaster\_IP -s 10.100.2.34 -j ACCEPT\\
-      \sout{{\color{red}-A webmaster\_IP -j DROP}}\\
-      {\color{blue}-A webmaster\_IP -j admin\_IP}}
-    \end{example}
-    Jump to the sysadmin chain at the end of the webmaster chain.\\
-    \textbf{Warning :} check conditions of ending chains
-  \end{frame}
-  \begin{frame}
-    \frametitle{Tutorial use case : access to SMB}
-    The rules for SMB (tcp port 139 and 445) can be reduced to :
-    \begin{example}
-      \small{\sout{{\color{red}-A INPUT -p tcp -m tcp \dd dport 445 -j admin\_IP}}\\
-      -A INPUT -p tcp -m tcp \dd dport 445 -j webmaster\_IP\\
-      \sout{{\color{red}-A INPUT -p tcp -m tcp \dd dport 139 -j admin\_IP}}\\
-      -A INPUT -p tcp -m tcp \dd dport 139 -j webmaster\_IP}
-    \end{example}
-  The admin chain is checked as well because the webmaster chain includes it.
   \end{frame}
@@
   \end{frame}
-  \begin{frame}
-    \frametitle{IPv6}
-    Configuration is the same for IPv4, use \textit{ip6tables} :
-    \begin{example}
-      \small{ip6tables -P INPUT DROP\\
-      ip6tables -P OUTPUT ACCEPT\\
-      ip6tables -A INPUT -i lo -j ACCEPT \# allow local traffic\\
-      ip6tables -A INPUT -p icmp -j ACCEPT \# allow ping, etc\\
-      ip6tables -A INPUT -m state \dd state ESTABLISHED -j ACCEPT\\
-      ip6tables -A INPUT -m state \dd state RELATED -j ACCEPT}
-    \end{example}
-    Even if IPv4 rules are set, IPv6 has to be configured seperately, otherwise the defaults apply.\\
-    \pause
-    \textbf{IPv6 day (June 6th 2014)} $\rightarrow$ disable IPv4 for one day
-  \end{frame}
 \end{document}

### 2015-03-23 buildtimetrend/presentation.tex
update title and links

@@
 
-\title[Buildtime Trend]{Buildtime Trend}
+\title[Buildtime Trend : What, why and how]{Buildtime Trend : What's trending in your build process}
 \author{Dieter Adriaenssens}
-\institute[Buildtime Trend]{@dcadriaenssens - Buildtime Trend}
+\institute[Buildtime Trend]{@dcadriaenssens - Founder, developer Buildtime Trend}
 \date[NewLine 5Apr2015]{NewLine - Ghent\\
@@
     \begin{itemize}
-      \item Brief introduction to iptables/netfilter
-      \item Optimizing configuration with a tutorial use case
-      \item Conclusion
+      \item What is Buildtime Trend
+      \item Why and how
+      \item Demo
+      \item Lessons learned
     \end{itemize}
@@
   \begin{frame}
-    \frametitle{What is iptables/netfilter?}
+    \frametitle{What is Buildtime Trend?}
     %(question to audience : who knows/uses iptables?)
-    Iptables is a tool for creating the rulesets for netfilter, a packet filtering framework which was introduced in the linux 2.4 kernel
   \end{frame}
-
   \begin{frame}
-    \frametitle{Conclusion}
+    \frametitle{Lessons learned}
     \begin{itemize}
-      \item use ESTABLISHED state to reduce number of rules
-      \item using chains makes your rules easier readable and maintainable
-      \item chains can be reused for several rules
-      \item chains can be chained together
-      \item faster, because it only jumps to a chain when a rule matches
+      \item keep it simple
     \end{itemize}
@@
     \begin{itemize}
-      \item Blog : \href{http://ruleant.blogspot.com/}{http://ruleant.blogspot.com}
-      \item Twitter : \href{https://twitter.com/dcadriaenssens}{@dcadriaenssens}
+      \item Website : \href{https://buildtimetrend.github.io/}{https://buildtimetrend.github.io/}
+      \item Twitter : \href{https://twitter.com/buildtime_trend}{@buildtime\_trend}
     \end{itemize}

### 2015-03-23 buildtimetrend/presentation.tex
add logo
change affiliation
add slides about start of the project

@@
 
+\logo{
+  \includegraphics[scale=.1]{pictures/logo_buildtimetrend.png}
+}
 \title[Buildtime Trend : What, why and how]{Buildtime Trend : What's trending in your build process}
 \author{Dieter Adriaenssens}
-\institute[Buildtime Trend]{@dcadriaenssens - Founder, developer Buildtime Trend}
+\institute[Buildtime Trend]{Buildtime Trend founder, developer - @dcadriaenssens}
 \date[NewLine 5Apr2015]{NewLine - Ghent\\
@@
   \begin{frame}
+    \frametitle{It started with an itch}
+    \begin{itemize}
+      \item Some builds took longer than others
+      \item no timing information was present in the logs
+      \item which stage took longer?
+    \end{itemize}
+    \pause
+    \textbf{Solution:} create a script to generate timestamps
+  \end{frame}
+  \begin{frame}
+    \frametitle{Next step : calculate duration and generate chart}
+    Requirements :
+    \begin{itemize}
+      \item script to parse the generated timestamps
+      \item calculate duration of the stages
+      \item store the timing data of every build
+      \item use this data to generate the chart
+    \end{itemize}
+    \pause
+    One solution : bash + CSV + gnuplot\\
+    \pause
+    Another solution : Python + XML + matplotlib
+  \end{frame}
+  \begin{frame}
+    \frametitle{Problem}
+    I hadn't used Python before.\\
+    \pause
+    A good opportunity to learn Python!
+  \end{frame}
+  \begin{frame}
+    \frametitle{Learning Python}
+    \begin{itemize}
+      \item start with a tutorial
+      \item read documentation
+      \item ask Google and Stack Overflow
+      \item talk to a friend or colleague
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Other helpful things}
+    \begin{itemize}
+      \item check coding style
+        \begin{itemize}
+          \item commandline : pep8, pylint
+          \item online tools : Landscape.io, Scrutinizer
+        \end{itemize}
+      \item unit testing and test driven development (TDD)
+      \item code coverage
+      \item automate this with Continuous Integration (CI) : Travis CI
+      \item version control : Git, GitHub, ...
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Proof of concept}
+    Collection of Bash and Python scripts, generating, analysing and visualising timing data:
+  \end{frame}
+
+  \begin{frame}
     \frametitle{Lessons learned}
@@
       \item keep it simple
+      \item (re)use as much existing tools as possible
     \end{itemize}

### 2015-03-24 buildtimetrend/presentation.tex
improve wording
add more slides

@@
       \item ask Google and Stack Overflow
-      \item talk to a friend or colleague
+      \item get advice from a friend or colleague
+      \item go to talks and conferences
     \end{itemize}
@@
     Collection of Bash and Python scripts, generating, analysing and visualising timing data:
+    <show picture of matplotlib chart>
+  \end{frame}
+  \begin{frame}
+    \frametitle{Limitations}
+    \begin{itemize}
+      \item scalability of XML
+      \item querying data is limited
+      \item developing new charts is not efficient
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Integration with Travis CI}
+    Call Buildtime Trend scripts in .travis.yml file
+    \begin{itemize}
+      \item generate timestamps
+      \item install dependencies
+      \item analyze timestamps and generate chart
+      \item store timing data (XML) and chart in gh-pages
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Integration with Travis CI}
+    This is not ideal!
+    \begin{itemize}
+      \item scripts become part of the build process
+      \item build duration is influenced
+      \item complicated to set up
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Integration with Travis CI}
+    Possible solutions
+    \begin{itemize}
+      \item create a service that analyses the logfile after the build is finished
+      \item store the data in a database outside Travis
+      \item host the generated chart(s) outside Travis
+    \end{itemize}
+    But how?
+  \end{frame}
+  \begin{frame}
+    \frametitle{Keen.io}
+    <insert Keen.io logo>
+    Keen IO's powerful APIs do the heavy lifting for you, so you can gather all the data you want and start getting the answers you need. 
+  \end{frame}
+  \begin{frame}
+    \frametitle{Keen.io}
+    \begin{itemize}
+      \item gather data
+      \item store it
+      \item analyse
+      \item visualise
+      \item support for several languages, including Python
+    \end{itemize}
+    Great : this solves the data storage and chart generation problem!
+  \end{frame}
+  \begin{frame}
+    \frametitle{First release}
+    Dashboard :\\
+    <show picture of first dashboard>
+  \end{frame}
+  \begin{frame}
+    \frametitle{First release}
+    \begin{itemize}
+      \item scripts tightly coupled with Travis CI build
+      \item storing in XML and chart generation with matplotlib (native mode)
+      \item storing and chart generation with Keen.io
+      \item chart dashboard deployed to gh-pages
+    \end{itemize}
+   Yeah, it works!\\
+    \pause
+    But there is still room for improvement.
+  \end{frame}
+  \begin{frame}
+    \frametitle{Travis CI timing data}
+    Good news!\\
+    Travis CI announces adding timing data to their logs!\\
+  \end{frame}
+  \begin{frame}
+    \frametitle{Create a service}
+    \begin{itemize}
+      \item get Travis CI build logfile
+      \item parse the timing data
+      \item store data in Keen.io
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Rearrange project structure}
+    Create organisation on GitHub to host the different repositories :
+    \begin{itemize}
+      \item service
+      \item client
+      \item library (with common functionality)
+      \item dashboard
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Challenges}
+    \begin{itemize}
+      \item hosting the service
+      \item triggering the service at the end of a build
+      \item let the subprojects work together
+      \item Open Source license
+      \item business model
+    \end{itemize}
   \end{frame}
@@
     \begin{itemize}
-      \item keep it simple
+      \item start simple
       \item (re)use as much existing tools as possible
+      \item start with what you know, learn new skills as you progress
+      \item try, even if it is all scary at first
     \end{itemize}
@@
     \vfill
-    Contact
+    Buildtime Trend
     \begin{itemize}

### 2015-03-28 buildtimetrend/presentation.tex
add author name
add slides about heroku service

@@
   \end{frame}
+  \begin{frame}
+    \frametitle{Hosting the service}
+    \begin{itemize}
+      \item CherryPy turns scripts into webservice
+        \begin{itemize}
+          \item dashboard
+          \item badge
+          \item parse buildlog
+        \end{itemize}
+      \item service hosted on Heroku : \href{https://buildtimetrend.herokuapp.com/}{https://buildtimetrend.herokuapp.com/}
+      \item deploy to Heroku with Travis CI build
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Travis CI triggers service}
+    Trigger the service at the end of a Travis CI build in .travis.yml :
+    \begin{example}
+      \small{notifications:\\
+      \ \ webhooks:\\
+      \ \ \ \ - https://buildtimetrend.herokuapp.com/travis}
+    \end{example}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Travis CI triggers service}
+    \begin{itemize}
+      \item A payload with build data is sent
+      \item service identifies repo and build ID
+      \item build job data and logfiles are retrieved from Travis CI
+      \item build job logfiles are parsed looking for timing tags
+      \item build stages duration is calculated
+      \item build job data is sent to Keen.io
+    \end{itemize}
+  \end{frame}
 
@@
     \vfill
+    Dieter Adriaenssens - \href{https://twitter.com/dcadriaenssens}{@dcadriaenssens}
+    \vfill
     Buildtime Trend

### 2015-03-29 buildtimetrend/presentation.tex
update event title
small improvements
add slide about deploying to Heroku
format some things in italics

@@
 \institute[Buildtime Trend]{Buildtime Trend founder, developer - @dcadriaenssens}
-\date[NewLine 5Apr2015]{NewLine - Ghent\\
+\date[NewLine 0x05 5Apr2015]{NewLine 0x05 - Ghent\\
 April 5th, 2015}
@@
     \begin{itemize}
-      \item script to parse the generated timestamps
+      \item parse the generated timestamps
       \item calculate duration of the stages
       \item store the timing data of every build
-      \item use this data to generate the chart
+      \item generate the chart
     \end{itemize}
@@
   \begin{frame}
-    \frametitle{First release}
+    \frametitle{First release - features}
     \begin{itemize}
@@
     \end{itemize}
+    <Add project schema>
   \end{frame}
@@
   \begin{frame}
+    \frametitle{Deploy to Heroku during Travis CI build}
+    In \textit{.travis.yml} :
+    \begin{example}
+      \small{deploy:\\
+      \ \ provider: heroku\\
+      \ \ api\_key:\\
+      \ \ \ \ secure: <secure key>\\
+      \ \ app:\\
+      \ \ \ \ deploy-prod: buildtimetrend}
+    \end{example}
+    Branch \textit{deploy-prod} is deployed to app \href{https://buildtimetrend.herokuapp.com/}{https://buildtimetrend.herokuapp.com/}
+  \end{frame}
+  \begin{frame}
     \frametitle{Travis CI triggers service}
-    Trigger the service at the end of a Travis CI build in .travis.yml :
+    Trigger the service at the end of a Travis CI build in \textit{.travis.yml} :
     \begin{example}

### 2015-03-29 buildtimetrend/presentation.tex
mention JSON
add demo

@@
     \begin{itemize}
-      \item A payload with build data is sent
+      \item A JSON payload with build data is sent
       \item service identifies repo and build ID
@@
       \item build job data is sent to Keen.io
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Demo}
+    \begin{itemize}
+      \item Service : \href{https://buildtimetrend.herokuapp.com/}{\small{https://buildtimetrend.herokuapp.com/}}
+      \item Dashboard : \href{https://buildtimetrend.herokuapp.com/dashboard/buildtimetrend/python-lib/index.html}{\small{https://buildtimetrend.herokuapp.com/dashboard/buildtimetrend/python-lib}}
+      \item Badges : \href{https://github.com/buildtimetrend/service\#badge-examples}{\small{https://github.com/buildtimetrend/service\#badge-examples}}
     \end{itemize}

### 2015-03-31 buildtimetrend/presentation.tex
add example of matplotlib chart
add dashboard 0.1 example
add keen.io logo
improve text

@@
     Collection of Bash and Python scripts, generating, analysing and visualising timing data:
-    <show picture of matplotlib chart>
+    \includegraphics[scale=.45]{pictures/example_matplotlib_trend.png}
   \end{frame}
@@
     \frametitle{Keen.io}
-    <insert Keen.io logo>
+    \includegraphics[scale=.20]{pictures/keenio_workflow.png}\\
     Keen IO's powerful APIs do the heavy lifting for you, so you can gather all the data you want and start getting the answers you need. 
@@
   \begin{frame}
-    \frametitle{First release}
-    Dashboard :\\
-    <show picture of first dashboard>
+    \frametitle{First release - dashboard}
+    \includegraphics[scale=.45]{pictures/example_dashboard.png}
   \end{frame}
@@
       \item scripts tightly coupled with Travis CI build
-      \item storing in XML and chart generation with matplotlib (native mode)
-      \item storing and chart generation with Keen.io
+      \item storing data in XML and chart generation with matplotlib (native mode)
+      \item storing data and chart generation with Keen.io
       \item chart dashboard deployed to gh-pages

### 2015-03-31 buildtimetrend/presentation.tex
add project structure diagram
set graphics path
add service link, change link layout

@@
 %% Enable license helpers
-\graphicspath{{../cc_beamer/}}
+\graphicspath{ {../cc_beamer/}{pictures/} }
 \input{../cc_beamer/cc_beamer}
@@
 \logo{
-  \includegraphics[scale=.1]{pictures/logo_buildtimetrend.png}
+  \includegraphics[scale=.1]{logo_buildtimetrend.png}
 }
@@
     Collection of Bash and Python scripts, generating, analysing and visualising timing data:
-    \includegraphics[scale=.45]{pictures/example_matplotlib_trend.png}
+    \includegraphics[scale=.45]{example_matplotlib_trend.png}
   \end{frame}
@@
     \frametitle{Keen.io}
-    \includegraphics[scale=.20]{pictures/keenio_workflow.png}\\
+    \includegraphics[scale=.20]{keenio_workflow.png}\\
     Keen IO's powerful APIs do the heavy lifting for you, so you can gather all the data you want and start getting the answers you need. 
@@
     \frametitle{First release - dashboard}
-    \includegraphics[scale=.45]{pictures/example_dashboard.png}
+    \includegraphics[scale=.45]{example_dashboard.png}
   \end{frame}
@@
     \frametitle{Rearrange project structure}
-    Create organisation on GitHub to host the different repositories :
-    \begin{itemize}
-      \item service
-      \item client
-      \item library (with common functionality)
-      \item dashboard
-    \end{itemize}
-    <Add project schema>
+    Create organisation on GitHub to host the different repositories :\\
+    \includegraphics[scale=.5]{buildtimetrend_diagram.png}
   \end{frame}
@@
     \vfill
-    Dieter Adriaenssens - \href{https://twitter.com/dcadriaenssens}{@dcadriaenssens}
+    Dieter Adriaenssens - \href{https://twitter.com/dcadriaenssens}{\small{@dcadriaenssens}}
     \vfill
@@
     \begin{itemize}
-      \item Website : \href{https://buildtimetrend.github.io/}{https://buildtimetrend.github.io/}
-      \item Twitter : \href{https://twitter.com/buildtime_trend}{@buildtime\_trend}
+      \item Website : \href{https://buildtimetrend.github.io/}{\small{https://buildtimetrend.github.io/}}
+      \item Service : \href{https://buildtimetrend.herokuapp.com/}{\small{https://buildtimetrend.herokuapp.com/}}
+      \item Twitter : \href{https://twitter.com/buildtime_trend}{\small{@buildtime\_trend}}
     \end{itemize}

### 2015-04-01 buildtimetrend/presentation.tex
remove unused macro
improve affiliation
add subtitle
improve text
add project description

@@
 \usetheme{Madrid}
-
-% macro for double dash
-\newcommand{\dd}{{\texttt{-{}-}}}
 
@@
 \title[Buildtime Trend : What, why and how]{Buildtime Trend : What's trending in your build process}
+\subtitle{What, why and how}
 \author{Dieter Adriaenssens}
-\institute[Buildtime Trend]{Buildtime Trend founder, developer - @dcadriaenssens}
+\institute[]{Buildtime Trend founder \& developer - @dcadriaenssens}
 \date[NewLine 0x05 5Apr2015]{NewLine 0x05 - Ghent\\
@@
     \frametitle{What is Buildtime Trend?}
-    %(question to audience : who knows/uses iptables?)
+    Buildtime Trend is an Open Source application that uses (timing) data to visualise trends in a build process.
   \end{frame}
@@
     \frametitle{It started with an itch}
+    I was working on a project that was built on Travis CI:
     \begin{itemize}
-      \item Some builds took longer than others
+      \item some builds took longer than others
       \item no timing information was present in the logs
@@
     \pause
-    One solution : bash + CSV + gnuplot\\
+    One solution : Bash + CSV + gnuplot\\
     \pause
@@
     \frametitle{Problem}
-    I hadn't used Python before.\\
+    I didn't use Python before.\\
     \pause

### 2015-04-02 buildtimetrend/presentation.tex
add slides about combining subprojects and Open Source license
add slide about Business plan

@@
     \begin{itemize}
+      \item let the subprojects work together
+      \item Open Source license
       \item hosting the service
       \item triggering the service at the end of a build
-      \item let the subprojects work together
-      \item Open Source license
       \item business model
     \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Combine subprojects}
+    \begin{itemize}
+    \item Client and service use dashboard : git submodule
+    \item Client and service use library :
+      \begin{itemize}
+        \item create Python package \textit{buildtimetrend}
+        \item available on \textit{pypi}
+        \item install as dependency
+      \end{itemize}
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Open Source license}
+    Transition from GNU General Public License (GPL) v3 to Affero GPLv3
   \end{frame}
@@
     \begin{itemize}
-      \item A JSON payload with build data is sent
+      \item a JSON payload with build data is sent
       \item service identifies repo and build ID
@@
       \item build job data is sent to Keen.io
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Business model}
+    \begin{itemize}
+      \item offering the service costs money
+        \begin{itemize}
+	  \item hosting the service
+          \item storing the data
+        \end{itemize}
+      \item Github inspired business model : free for Open Source, paying for private repos
+      \item \href{https://keen.io}{Keen.io} offered to host data for free for Open Source projects. \textbf{Thanks, guys!}
+      \item Further development with real data gathering : better feedback
     \end{itemize}

### 2015-04-02 buildtimetrend/presentation.tex
add slides about future and welcoming contributors
add acknowledgments

@@
   \begin{frame}
+    \frametitle{Future development}
+    \begin{itemize}
+      \item caching for badges and dashboard charts
+      \item service : seperate worker for build job processing
+      \item support private Github repos
+      \item support other CI platforms (Jenkins, ...)
+      \item more and improved metrics and trends
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Contributions welcome}
+    \begin{itemize}
+      \item use and test the service
+      \item report bugs
+      \item suggest improvements
+      \item clone the project and send pull requests
+      \begin{itemize}
+        \item implement a feature
+        \item fix a bug
+        \item add a new chart
+        \item improve dashboard layout
+        \item \ldots
+      \end{itemize}
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
     \frametitle{Demo}
@@
   \end{frame}
-
   \begin{frame}
@@
       \item try, even if it is all scary at first
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Acknowledgements}
+    Big thanks to
+    \begin{itemize}
+      \item Josh (\href{https://twitter.com/dzello}{@dzello}) and the other nice people of Keen.io, for their invaluable support!
+      \item \href{https://twitter.com/thomasbonte}{@thomasbonte} for introducing me to Lean Startup principles
+      \item Alex for telling me about CherryPy
+      \item all the \href{https://github.com/buildtimetrend/python-lib/wiki/Credits}{services} that power the project
+      \item my friends and family
     \end{itemize}

### 2015-04-04 oss365/presentation.tex
start oss365 presentation

\documentclass[14pt]{beamer}
\usepackage{hyperref}
\usepackage{ulem}
\usepackage[T1]{fontenc}
\usetheme{Madrid}

%% Enable license helpers
\graphicspath{ {../cc_beamer/}{pictures/} }
\input{../cc_beamer/cc_beamer}

\title[365 days of Open Source]{What I learned from 365 days of contributing to Open Source software}
\author{Dieter Adriaenssens}
\institute[]{@dcadriaenssens}
\date[NewLine 0x05 5Apr2015]{NewLine 0x05 - Ghent\\
April 5th, 2015}
\subject{365 days of Open Source}
\begin{document}
  \begin{frame}
    \titlepage
    \vfill
    \begin{center}
      \CcGroupByNcSa{0.83}{0.95ex}\\[2.5ex]
        {\tiny\CcNote{\CcLongnameByNcSa}}
        \vspace*{-2.5ex}
    \end{center}
  \end{frame}
  \begin{frame}
    \frametitle{365 day of Open Source}
    \begin{itemize}
      \item Github : 365 day commit streak, 2332 commits
      \item \href{https://github.com/ruleant}{Today : 447 day commit streak} (and counting)
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{How it started}
    \begin{itemize}
      \item 14 January 2014 : returned from 'unplugged' holiday
      \item previous 100+ day commit streaks
      \item see how far I would get
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{June 2014 : Getting serious}
    \begin{itemize}
      \item Josh (@dzello) of Keen.io pledged to do a 365 day commit streak
      \item I was at 162 days 
      \item I took the challenge to do a 365 day streak
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{How}
    \begin{itemize}
      \item working on my projects : Buildtime Trend, GetBack GPS
      \item new features 
      \item refactoring
      \item add unit tests
      \item coding style
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{Lessons learned}
    \begin{itemize}
      \item void
    \end{itemize}
  \end{frame}
  \begin{frame}
    \frametitle{Acknowledgements}
    Big thanks to
    \begin{itemize}
      \item Josh (\href{https://twitter.com/dzello}{@dzello}) of Keen.io
      \item my friends and family
    \end{itemize}
  \end{frame}
  \begin{frame}
   \frametitle{Questions}
    Thanks for your attention!\\
    Questions?
    \vfill
    Dieter Adriaenssens - \href{https://twitter.com/dcadriaenssens}{\small{@dcadriaenssens}}
  \end{frame}
\end{document}

### 2015-04-04 oss365/presentation.tex
add a few more slides
add more slides

@@
   \begin{frame}
+    \frametitle{Holiday}
+    \begin{itemize}
+      \item July 2014 : climbing trip
+      \item climbing during the day
+      \item limited time in the evening
+      \item planning to do small commits
+      \item close call on a few days
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Difficulties}
+    \begin{itemize}
+      \item Traveling
+      \item Tired after a long day at work
+      \item stuck on a problem
+    \end{itemize}
+    Will I keep my streak going?
+  \end{frame}
+  \begin{frame}
+    \frametitle{Meeting Josh}
+    \begin{itemize}
+      \item Meeting Josh in San Francisco in October 2014
+      \item Josh stopped his 66 day streak going to Burning Man
+      \item other ways of measuring progress
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
     \frametitle{Lessons learned}
     \begin{itemize}
-      \item void
+      \item set a realistic goal
+      \item choose something that works for you
+      \item minimum 1 commit per day worked for me
+      \item I will keep the streak going as long as I am comfortable with
+      \item working on the project should be the primary goal
+      \item pledge to commit every day is means to keep you going
+      \item or any sort of metrics
+      \item similarities in sports, goals, projects, \ldots
+      \item allow to give yourself a break
     \end{itemize}
@@
     Dieter Adriaenssens - \href{https://twitter.com/dcadriaenssens}{\small{@dcadriaenssens}}
+    \vfill
+    Blogpost : \href{http://ruleant.blogspot.be/2015/01/what-i-learned-from-365-days-of.html}{http://ruleant.blogspot.be/2015/01/what-i-learned-from-365-days-of.html}
   \end{frame}

### 2015-04-05 oss365/presentation.tex
add github streak snapshot
add picture of meeting Josh
add Open Source titles
add climbing picture
split conclusion
fix title

@@
 
-\title[365 days of Open Source]{What I learned from 365 days of contributing to Open Source software}
+\title[365 days of Open Source]{What I learned from 365 days of contributing to Open Source projects}
 \author{Dieter Adriaenssens}
-\institute[]{@dcadriaenssens}
+\institute[]{Open Source developer - @dcadriaenssens}
 \date[NewLine 0x05 5Apr2015]{NewLine 0x05 - Ghent\\
@@
     \frametitle{365 day of Open Source}
+    \includegraphics[scale=.45]{github_oss365_13jan2015.png}
     \begin{itemize}
@@
     \begin{itemize}
-      \item Josh (@dzello) of Keen.io pledged to do a 365 day commit streak
+      \item Josh (@dzello), Open Sourcerer at Keen.io pledged to do a 365 day commit streak
       \item I was at 162 days 
@@
   \begin{frame}
-    \frametitle{Meeting Josh}
+    \frametitle{Meeting Josh in San Francisco (Oct 2014)}
+    \includegraphics[scale=.1]{josh_justin.jpg}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Josh's experience}
     \begin{itemize}
-      \item Meeting Josh in San Francisco in October 2014
       \item Josh stopped his 66 day streak going to Burning Man
@@
       \item I will keep the streak going as long as I am comfortable with
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Lessons learned - continued}
+    \begin{itemize}
       \item working on the project should be the primary goal
       \item pledge to commit every day is means to keep you going
-      \item or any sort of metrics
+      \item or any sort of metric
       \item similarities in sports, goals, projects, \ldots
@@
     \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Break}
+    \includegraphics[scale=.25]{climbing.jpg}
   \end{frame}

### 2015-04-05 buildtimetrend/presentation.tex
mention first adopters
add more advice
some improvements
add link to presentation location

@@
       \item ask Google and Stack Overflow
-      \item get advice from a friend or colleague
+      \item read code from other projects
+      \item get advice from friends and colleagues
       \item go to talks and conferences
@@
           \item commandline : pep8, pylint
-          \item online tools : Landscape.io, Scrutinizer
+          \item online tools : Landscape.io, Scrutinizer, Codacy
         \end{itemize}
@@
       \item Alex for telling me about CherryPy
+      \item the Open Source projects testdriving the service
       \item all the \href{https://github.com/buildtimetrend/python-lib/wiki/Credits}{services} that power the project
@@
     \vfill
+    Presentation available on \href{https://ruleant.github.io/presentations/}{\small{https://ruleant.github.io/presentations/}}
+    \vfill
     Buildtime Trend

### 2016-01-26 buildtimetrend/btt_fosdem2016.tex
start working on Fosdem presentation

--- 2015-04-05 buildtimetrend/presentation.tex
@@
 }
-\title[Buildtime Trend : What, why and how]{Buildtime Trend : What's trending in your build process}
-\subtitle{What, why and how}
+\title[Buildtime Trend]{Buildtime Trend}
+\subtitle{What's trending in your build process}
 \author{Dieter Adriaenssens}
 \institute[]{Buildtime Trend founder \& developer - @dcadriaenssens}
-\date[NewLine 0x05 5Apr2015]{NewLine 0x05 - Ghent\\
-April 5th, 2015}
+\date[FOSDEM 2016]{FOSDEM 2016 - Brussels\\
+January 31st, 2016}
 \subject{Buildtime Trend}
@@
     \begin{itemize}
-      \item What is Buildtime Trend
-      \item Why and how
+      \item What is Buildtime Trend?
+      \item How does it work?
+      \item How to start using it
       \item Demo
-      \item Lessons learned
     \end{itemize}
@@
     \end{itemize}
-    \pause
-    \textbf{Solution:} create a script to generate timestamps
-  \end{frame}
-  \begin{frame}
-    \frametitle{Next step : calculate duration and generate chart}
-    Requirements :
-    \begin{itemize}
-      \item parse the generated timestamps
-      \item calculate duration of the stages
-      \item store the timing data of every build
-      \item generate the chart
-    \end{itemize}
-    \pause
-    One solution : Bash + CSV + gnuplot\\
-    \pause
-    Another solution : Python + XML + matplotlib
-  \end{frame}
-  \begin{frame}
-    \frametitle{Problem}
-    I didn't use Python before.\\
-    \pause
-    A good opportunity to learn Python!
-  \end{frame}
-  \begin{frame}
-    \frametitle{Learning Python}
-    \begin{itemize}
-      \item start with a tutorial
-      \item read documentation
-      \item ask Google and Stack Overflow
-      \item read code from other projects
-      \item get advice from friends and colleagues
-      \item go to talks and conferences
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Other helpful things}
-    \begin{itemize}
-      \item check coding style
-        \begin{itemize}
-          \item commandline : pep8, pylint
-          \item online tools : Landscape.io, Scrutinizer, Codacy
-        \end{itemize}
-      \item unit testing and test driven development (TDD)
-      \item code coverage
-      \item automate this with Continuous Integration (CI) : Travis CI
-      \item version control : Git, GitHub, ...
-    \end{itemize}
   \end{frame}
@@
   \begin{frame}
-    \frametitle{Limitations}
+    \frametitle{How it works}
+    Service :
     \begin{itemize}
-      \item scalability of XML
-      \item querying data is limited
-      \item developing new charts is not efficient
+      \item get Travis CI build logfile
+      \item parse the (timing) data
+      \item store data
+      \item visualise using a dashboard with trends
     \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Integration with Travis CI}
-    Call Buildtime Trend scripts in .travis.yml file
-    \begin{itemize}
-      \item generate timestamps
-      \item install dependencies
-      \item analyze timestamps and generate chart
-      \item store timing data (XML) and chart in gh-pages
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Integration with Travis CI}
-    This is not ideal!
-    \begin{itemize}
-      \item scripts become part of the build process
-      \item build duration is influenced
-      \item complicated to set up
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Integration with Travis CI}
-    Possible solutions
-    \begin{itemize}
-      \item create a service that analyses the logfile after the build is finished
-      \item store the data in a database outside Travis
-      \item host the generated chart(s) outside Travis
-    \end{itemize}
-    But how?
   \end{frame}
@@
   \begin{frame}
-    \frametitle{Keen.io}
-    \begin{itemize}
-      \item gather data
-      \item store it
-      \item analyse
-      \item visualise
-      \item support for several languages, including Python
-    \end{itemize}
-    Great : this solves the data storage and chart generation problem!
-  \end{frame}
-  \begin{frame}
     \frametitle{First release - dashboard}
     \includegraphics[scale=.45]{example_dashboard.png}
-  \end{frame}
-  \begin{frame}
-    \frametitle{First release - features}
-    \begin{itemize}
-      \item scripts tightly coupled with Travis CI build
-      \item storing data in XML and chart generation with matplotlib (native mode)
-      \item storing data and chart generation with Keen.io
-      \item chart dashboard deployed to gh-pages
-    \end{itemize}
-    Yeah, it works!\\
-    \pause
-    But there is still room for improvement.
-  \end{frame}
-  \begin{frame}
-    \frametitle{Travis CI timing data}
-    Good news!\\
-    Travis CI announces adding timing data to their logs!\\
-  \end{frame}
-  \begin{frame}
-    \frametitle{Create a service}
-    \begin{itemize}
-      \item get Travis CI build logfile
-      \item parse the timing data
-      \item store data in Keen.io
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Rearrange project structure}
-    Create organisation on GitHub to host the different repositories :\\
-    \includegraphics[scale=.5]{buildtimetrend_diagram.png}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Challenges}
-    \begin{itemize}
-      \item let the subprojects work together
-      \item Open Source license
-      \item hosting the service
-      \item triggering the service at the end of a build
-      \item business model
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Combine subprojects}
-    \begin{itemize}
-    \item Client and service use dashboard : git submodule
-    \item Client and service use library :
-      \begin{itemize}
-        \item create Python package \textit{buildtimetrend}
-        \item available on \textit{pypi}
-        \item install as dependency
-      \end{itemize}
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Open Source license}
-    Transition from GNU General Public License (GPL) v3 to Affero GPLv3
   \end{frame}
@@
           \item dashboard
-          \item badge
+          \item badges
           \item parse buildlog
@@
       \item service hosted on Heroku : \href{https://buildtimetrend.herokuapp.com/}{https://buildtimetrend.herokuapp.com/}
-      \item deploy to Heroku with Travis CI build
     \end{itemize}
@@
   \begin{frame}
-    \frametitle{Deploy to Heroku during Travis CI build}
-    In \textit{.travis.yml} :
-    \begin{example}
-      \small{deploy:\\
-      \ \ provider: heroku\\
-      \ \ api\_key:\\
-      \ \ \ \ secure: <secure key>\\
-      \ \ app:\\
-      \ \ \ \ deploy-prod: buildtimetrend}
-    \end{example}
-    Branch \textit{deploy-prod} is deployed to app \href{https://buildtimetrend.herokuapp.com/}{https://buildtimetrend.herokuapp.com/}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Travis CI triggers service}
+    \frametitle{How to use it}
     Trigger the service at the end of a Travis CI build in \textit{.travis.yml} :
@@
     \end{example}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Travis CI triggers service}
-    \begin{itemize}
-      \item a JSON payload with build data is sent
-      \item service identifies repo and build ID
-      \item build job data and logfiles are retrieved from Travis CI
-      \item build job logfiles are parsed looking for timing tags
-      \item build stages duration is calculated
-      \item build job data is sent to Keen.io
-    \end{itemize}
   \end{frame}
@@
       \item caching for badges and dashboard charts
-      \item service : seperate worker for build job processing
       \item support private Github repos
@@
   \begin{frame}
-    \frametitle{Lessons learned}
-    \begin{itemize}
-      \item start simple
-      \item (re)use as much existing tools as possible
-      \item start with what you know, learn new skills as you progress
-      \item try, even if it is all scary at first
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
     \frametitle{Acknowledgements}
@@
     \begin{itemize}
-      \item Josh (\href{https://twitter.com/dzello}{@dzello}) and the other nice people of Keen.io, for their invaluable support!
-      \item \href{https://twitter.com/thomasbonte}{@thomasbonte} for introducing me to Lean Startup principles
-      \item Alex for telling me about CherryPy
+      \item The nice people of Keen.io, for their invaluable support!
       \item the Open Source projects testdriving the service
       \item all the \href{https://github.com/buildtimetrend/python-lib/wiki/Credits}{services} that power the project
-      \item my friends and family
     \end{itemize}

### 2016-01-28 buildtimetrend/btt_fosdem2016.tex
add slide with projects
some small improvements

@@
 \institute[]{Buildtime Trend founder \& developer - @dcadriaenssens}
-\date[FOSDEM 2016]{FOSDEM 2016 - Brussels\\
+\date[FOSDEM 2016]{FOSDEM 2016\\
 January 31st, 2016}
@@
   \begin{frame}
-    \frametitle{What is Buildtime Trend?}
-    Buildtime Trend is an Open Source application that uses (timing) data to visualise trends in a build process.
-  \end{frame}
-  \begin{frame}
     \frametitle{It started with an itch}
-    I was working on a project that was built on Travis CI:
+    Project with unreliable builds on Travis CI:
     \begin{itemize}
@@
     \includegraphics[scale=.45]{example_matplotlib_trend.png}
+  \end{frame}
+  \begin{frame}
+    \frametitle{What is Buildtime Trend?}
+    Buildtime Trend is an Open Source application that uses (timing) data to visualise trends in a build process
   \end{frame}
@@
       \item service hosted on Heroku : \href{https://buildtimetrend.herokuapp.com/}{https://buildtimetrend.herokuapp.com/}
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Who is using it}
+    \pause
+    \begin{itemize}
+      \item Angular
+      \item RocksDB (Facebook)
+      \item phpMyAdmin
+      \item Weblate
+      \item MuseScore
+      \item Servo
+      \item \ldots
     \end{itemize}

### 2016-01-28 buildtimetrend/btt_fosdem2016.tex
move dashboard screenshot forward, improve text
move who's using it, change 'free for open source' slide

@@
     \frametitle{Proof of concept}
-    Collection of Bash and Python scripts, generating, analysing and visualising timing data:
+    A script to generate a chart with duration of build stages :
     \includegraphics[scale=.45]{example_matplotlib_trend.png}
+  \end{frame}
+  \begin{frame}
+    \frametitle{First release - dashboard}
+    \includegraphics[scale=.45]{example_dashboard.png}
   \end{frame}
@@
     \frametitle{How it works}
-    Service :
+    Webservice is triggered at the end of a build :
     \begin{itemize}
-      \item get Travis CI build logfile
+      \item download Travis CI build logfile
       \item parse the (timing) data
       \item store data
-      \item visualise using a dashboard with trends
+      \item visualise using a dashboard with charts
     \end{itemize}
@@
   \begin{frame}
-    \frametitle{First release - dashboard}
-    \includegraphics[scale=.45]{example_dashboard.png}
-  \end{frame}
-  \begin{frame}
     \frametitle{Hosting the service}
@@
           \item badges
-          \item parse buildlog
+          \item buildlog parser
         \end{itemize}
       \item service hosted on Heroku : \href{https://buildtimetrend.herokuapp.com/}{https://buildtimetrend.herokuapp.com/}
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Who is using it}
-    \pause
-    \begin{itemize}
-      \item Angular
-      \item RocksDB (Facebook)
-      \item phpMyAdmin
-      \item Weblate
-      \item MuseScore
-      \item Servo
-      \item \ldots
     \end{itemize}
@@
   \begin{frame}
-    \frametitle{Business model}
+    \frametitle{Who can use it}
     \begin{itemize}
-      \item offering the service costs money
+      \item Webservice : Free for Open Source (for now)
+      \item \href{https://keen.io}{Keen.io} offered to host data for free for Open Source projects. \textbf{Thanks, guys!}
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{Who is using it}
     \begin{itemize}
-	  \item hosting the service
-          \item storing the data
-        \end{itemize}
-      \item Github inspired business model : free for Open Source, paying for private repos
-      \item \href{https://keen.io}{Keen.io} offered to host data for free for Open Source projects. \textbf{Thanks, guys!}
-      \item Further development with real data gathering : better feedback
+      \item Angular
+      \item RocksDB (Facebook)
+      \item phpMyAdmin
+      \item Weblate
+      \item MuseScore
+      \item Servo
+      \item pyca/cryptography
+      \item Pootle
+      \item \ldots
     \end{itemize}

### 2016-01-29 buildtimetrend/btt_fosdem2016.tex
update future roadmap
move description and demo slides

@@
   \begin{frame}
+    \frametitle{What is Buildtime Trend?}
+    Buildtime Trend is an Open Source application that uses (timing) data to visualise trends in a build process
+  \end{frame}
+  \begin{frame}
     \frametitle{It started with an itch}
@@
     \includegraphics[scale=.45]{example_dashboard.png}
-  \end{frame}
-  \begin{frame}
-    \frametitle{What is Buildtime Trend?}
-    Buildtime Trend is an Open Source application that uses (timing) data to visualise trends in a build process
   \end{frame}
@@
   \begin{frame}
+    \frametitle{Demo}
+    \begin{itemize}
+      \item Service : \href{https://buildtimetrend.herokuapp.com/}{\small{https://buildtimetrend.herokuapp.com/}}
+      \item Badges : \href{https://github.com/buildtimetrend/service\#badge-examples}{\small{https://github.com/buildtimetrend/service\#badge-examples}}
+    \end{itemize}
+  \end{frame}
+  \begin{frame}
     \frametitle{Future development}
     \begin{itemize}
-      \item caching for badges and dashboard charts
-      \item support private Github repos
+      \item more and improved metrics and charts
+      \item improve navigation and searching on dashboard
       \item support other CI platforms (Jenkins, ...)
-      \item more and improved metrics and trends
+      \item notification on a jumpy build job
     \end{itemize}
@@
       \end{itemize}
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{Demo}
-    \begin{itemize}
-      \item Service : \href{https://buildtimetrend.herokuapp.com/}{\small{https://buildtimetrend.herokuapp.com/}}
-      \item Dashboard : \href{https://buildtimetrend.herokuapp.com/dashboard/buildtimetrend/python-lib/index.html}{\small{https://buildtimetrend.herokuapp.com/dashboard/buildtimetrend/python-lib}}
-      \item Badges : \href{https://github.com/buildtimetrend/service\#badge-examples}{\small{https://github.com/buildtimetrend/service\#badge-examples}}
     \end{itemize}

### 2016-01-30 buildtimetrend/btt_fosdem2016.tex
improve text
better explain packages to be used

@@
           \item badges
-          \item buildlog parser
+          \item buildjob log parser
         \end{itemize}
@@
     \begin{itemize}
-      \item Webservice : Free for Open Source (for now)
-      \item \href{https://keen.io}{Keen.io} offered to host data for free for Open Source projects. \textbf{Thanks, guys!}
+      \item Webservice : free for Open Source (for now)
+      \begin{itemize}
+        \item \href{https://keen.io}{Keen.io} offered to host data for Open Source projects. \textbf{Thanks, guys!}
+      \end{itemize}
+      \item Source code available on Github
+      \begin{itemize}
+        \item service : host it yourself
+        \item client : implement in your build process
+      \end{itemize}
     \end{itemize}
@@
       \item support other CI platforms (Jenkins, ...)
-      \item notification on a jumpy build job
+      \item notifications
     \end{itemize}
@@
     Thanks for your attention!\\
-    Questions?
     \vfill
-    Dieter Adriaenssens - \href{https://twitter.com/dcadriaenssens}{\small{@dcadriaenssens}}
+    \href{https://twitter.com/dcadriaenssens}{\small{@dcadriaenssens}}
     \vfill

### 2016-06-19 oss365/presentation_newline2015.tex
rename oss365 newline presentation

--- 2015-04-05 oss365/presentation.tex
@@
     \begin{itemize}
-      \item 14 January 2014 : returned from 'unplugged' holiday
+      \item 14 January 2014 : I returned from an 'unplugged' holiday
       \item previous 100+ day commit streaks
@@
     \end{itemize}
+  \end{frame}
+  \begin{frame}
+    \frametitle{How}
+    \begin{itemize}
+      \item working on my projects : Buildtime Trend, GetBack GPS, phpMyAdmin
+      \item bigger tasks :
+      \begin{itemize}
+        \item new features
+        \item bug fixes
+        \item refactoring
+        \item add unit tests
+      \end{itemize}
+      \item smaller tasks :
+      \begin{itemize}
+        \item fix coding style
+        \item fix typos
+        \item translations
+      \end{itemize}
+    \end{itemize}
+    => code quality improves
   \end{frame}
@@
       \item I took the challenge to do a 365 day streak
-    \end{itemize}
-  \end{frame}
-  \begin{frame}
-    \frametitle{How}
-    \begin{itemize}
-      \item working on my projects : Buildtime Trend, GetBack GPS
-      \item new features 
-      \item refactoring
-      \item add unit tests
-      \item coding style
     \end{itemize}
